---
title: "Cleaning"
subtitle: "What was removed, and how duplicates and replicates were collapsed"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Step 2 removes chemistry measurements that failed quality control, then collapses
repeated measurements: identical values (duplicates) are kept once, and differing
values of the same measurand on the same occasion (technical replicates) are
averaged, keeping the mean, its standard deviation, and the replicate count.

```{r}
#| label: setup
library(DBI)
library(RSQLite)
library(tibble)
library(dplyr)
library(knitr)

db_dir <- "../data/db"
srcs <- tibble(
  Source = c("Mareano", "Vannmiljø", "ICES-DOME", "MUDAB", "4Demon"),
  stem   = c("mareano", "vannmiljo", "ices_dome", "mudab", "4demon"))
kfmt <- function(df, ...) kable(df, format.args = list(big.mark = ","), ...)

# Removal counts come from the slim DB (which still carries the flags), scoped to
# chemistry; src_flag is only present for some sources.
removal_counts <- function(stem) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(stem, "_slim.sqlite")))
  on.exit(dbDisconnect(con))
  has_src <- "src_flag" %in% dbListFields(con, "measurement")
  sql <- paste0(
    "SELECT COUNT(*) chem,
       SUM(range_flag IS NOT NULL) out_of_range,
       SUM(invalid_flag IS NOT NULL) invalid,
       SUM(below_loq=1 OR below_loq_num=1) below_limit,
       SUM(weight_basis='wet') wet",
    if (has_src) ", SUM(src_flag IS NOT NULL) source_qc" else ", 0 source_qc",
    " FROM measurement m JOIN element e ON m.symbol=e.symbol
      WHERE e.category IN ('target','reference','organic')")
  as_tibble(dbGetQuery(con, sql))
}

clean_stats <- function(stem) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(stem, "_clean.sqlite")))
  on.exit(dbDisconnect(con))
  as_tibble(dbGetQuery(con,
    "SELECT COUNT(*) kept, SUM(n_rep=1) single, SUM(n_rep>1) averaged,
            MAX(n_rep) max_rep, SUM(value_uncrt IS NOT NULL) with_uncrt
     FROM measurement"))
}
```

## 1. Rows removed by flag

Chemistry measurements carrying each flag (from the slim stage), scoped to
chemistry. A measurement can carry more than one flag, so the removed total is the
union, not the sum. Everything flagged here is dropped from the clean database.

```{r}
bind_rows(lapply(seq_len(nrow(srcs)), function(i)
  removal_counts(srcs$stem[i]) |> mutate(Source = srcs$Source[i]))) |>
  transmute(Source,
            `Chemistry (slim)` = chem,
            `Out of range` = out_of_range,
            `Invalid` = invalid,
            `Below limit` = below_limit,
            `Wet weight` = wet,
            `Source QC` = source_qc) |>
  kfmt()
```

## 2. Duplicates and technical replicates

After removal, the surviving chemistry is collapsed per sampling occasion + element
+ method. Most measurements are singletons (`n_rep = 1`); groups with more than one
distinct value are averaged (`n_rep > 1`), carrying a standard deviation.

```{r}
bind_rows(lapply(seq_len(nrow(srcs)), function(i)
  clean_stats(srcs$stem[i]) |> mutate(Source = srcs$Source[i]))) |>
  transmute(Source,
            `Clean rows` = kept,
            `Single` = single,
            `Averaged (replicates)` = averaged,
            `Max replicates` = max_rep) |>
  kfmt()
```

## 3. Measurement uncertainty

Where a source reports analytical uncertainty, it is standardised to mg/kg
(`value_uncrt`) alongside the value. Only ICES-DOME carries per-measurement
uncertainty (`uncrt` / `metcu`), so it is the only source with a populated
`value_uncrt`.

```{r}
bind_rows(lapply(seq_len(nrow(srcs)), function(i)
  clean_stats(srcs$stem[i]) |> mutate(Source = srcs$Source[i]))) |>
  transmute(Source, `Clean rows` = kept, `With uncertainty` = with_uncrt) |>
  kfmt()
```
