---
title: "Clean schema"
subtitle: "The tables and columns of the clean databases"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Each `\<source\>_clean.sqlite` has the seven shared tables carried from the slim
schema, plus two grain-size tables (absent for 4Demon, which has no grain-size).
The differences from the slim schema are that `measurement` is chemistry-only with
aggregated values, grain-size lives in its own tables, and depths are in cm.

```{r}
library(tibble)
library(knitr)
```

## measurement (chemistry only)

One row per cleaned chemistry measurement (target / reference / organic). Failed-QC
rows are gone; duplicates and technical replicates are collapsed.

```{r}
tribble(
  ~Column,        ~Type,    ~Meaning,
  "measurement_id","INTEGER","Primary key (carried from slim).",
  "subsample_id",  "INTEGER","-> subsample.",
  "symbol",        "TEXT",   "Element, ICES name (-> element).",
  "value",         "REAL",   "Value in the original (ICES-named) unit.",
  "unit",          "TEXT",   "Original reporting unit.",
  "value_std",     "REAL",   "Standardised value, mg/kg (mean where replicates were averaged).",
  "unit_std",      "TEXT",   "Always mg/kg.",
  "value_sd",      "REAL",   "SD across technical replicates, else NULL.",
  "n_rep",         "INTEGER","Number of distinct replicate values averaged (1 = single).",
  "value_uncrt",   "REAL",   "Analytical uncertainty, mg/kg (ICES-DOME only), else NULL.",
  "matrix",        "TEXT",   "Sample fraction the chemistry was measured on, where recorded.",
  "method_id",     "INTEGER","-> method."
) |> kable()
```

## grain_size (per subsample)

One row per subsample with usable grain-size.

```{r}
tribble(
  ~Column,       ~Type,     ~Meaning,
  "subsample_id","INTEGER", "Primary key (-> subsample).",
  "frac_class",  "TEXT",    "'bulk' (whole sample) or 'sieved' (a fine fraction).",
  "fines_lt63",  "REAL",    "% finer than 63 um (clay + silt), else NULL.",
  "is_fine",     "INTEGER", "1 if fines_lt63 >= 50% (mud-dominated), 0 if not, NULL if unknown.",
  "fines_basis", "TEXT",    "How fines_lt63 was derived."
) |> kable()
```

## grain_size_fraction (per fraction measurement)

The grain-size distribution moved out of `measurement`.

```{r}
tribble(
  ~Column,       ~Type,    ~Meaning,
  "subsample_id","INTEGER","-> subsample.",
  "symbol",      "TEXT",   "The source's grain-size code.",
  "matrix",      "TEXT",   "Sample fraction (where recorded).",
  "lo_um",       "REAL",   "Lower size bound, um.",
  "hi_um",       "REAL",   "Upper size bound, um (NULL = open, e.g. gravel).",
  "value_pct",   "REAL",   "Corrected mass fraction, %."
) |> kable()
```

## Unchanged tables

`element` (symbol, name, category), `dataset`, `site`, `event` (date, year, gear;
`datetime`/`time` dropped), `subsample` (depth in cm + supporting-data flags), and
`method` (analysis + `lod`/`loq` with `limit_unit`) carry over from slim with the
harmonisations applied.

## What changed from slim

```{r}
tribble(
  ~Change,                              ~Detail,
  "Rows removed",                       "Out of range, invalid, below limit, wet weight, source-specific QC.",
  "Duplicates / replicates collapsed",  "One value each; replicates gain value_sd + n_rep.",
  "measurement is chemistry only",      "Grain-size moved to grain_size / grain_size_fraction.",
  "Flag columns dropped",               "The slim QC flags were consumed by the cleaning and are not carried.",
  "Depths in cm",                       "ICES-DOME converted from metres; Vannmiljø corrupt depths nulled.",
  "Symbols / units unified",            "ICES names; TOC -> CORG; value_std in mg/kg added."
) |> kable()
```
