---
title: "Grain size"
subtitle: "Sieved/bulk, the <63 µm fines, and the fraction distribution"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Step 3 moves grain-size out of the measurement table into two tables: a
per-subsample **`grain_size`** summary (sieved or bulk, the `<63 µm` fines, and a
fine/not flag) and a **`grain_size_fraction`** detail table (the size
distribution). 4Demon has no grain-size and so does not appear here.

```{r}
#| label: setup
library(DBI)
library(RSQLite)
library(tibble)
library(dplyr)
library(tidyr)
library(knitr)

db_dir <- "../data/db"
gsrcs <- tibble(
  Source = c("Mareano", "Vannmiljø", "ICES-DOME", "MUDAB"),
  stem   = c("mareano", "vannmiljo", "ices_dome", "mudab"))

qry <- function(stem, sql) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(stem, "_clean.sqlite")))
  on.exit(dbDisconnect(con)); as_tibble(dbGetQuery(con, sql))
}
kfmt <- function(df, ...) kable(df, format.args = list(big.mark = ","), ...)
```

## 1. Sieved or bulk

Whether each grain-size sample is a whole (bulk) sediment or a sieved fraction.
For ICES-DOME and MUDAB this comes from the sample `matrix`; Mareano samples are
all bulk (confirmed); Vannmiljø has no sieving metadata and is taken as bulk.
Subsamples where this could not be determined were dropped.

```{r}
bind_rows(lapply(seq_len(nrow(gsrcs)), function(i)
  qry(gsrcs$stem[i], "SELECT frac_class, COUNT(*) n FROM grain_size GROUP BY frac_class") |>
    mutate(Source = gsrcs$Source[i]))) |>
  pivot_wider(names_from = frac_class, values_from = n, values_fill = 0) |>
  relocate(Source) |>
  kfmt()
```

## 2. The <63 µm fines

`fines_lt63` is the percentage of material finer than 63 µm (the clay + silt
"mud"). `is_fine` marks samples that are mud-dominated (at least 50 %). Samples
with a class but no usable fines value are `NA`.

```{r}
bind_rows(lapply(seq_len(nrow(gsrcs)), function(i)
  qry(gsrcs$stem[i], "SELECT
        COUNT(*) n_samples,
        SUM(is_fine=1) fine, SUM(is_fine=0) not_fine, SUM(is_fine IS NULL) unknown,
        ROUND(AVG(fines_lt63),1) mean_fines
      FROM grain_size") |>
    mutate(Source = gsrcs$Source[i]))) |>
  transmute(Source, `Grain-size samples` = n_samples,
            `Fine (>=50%)` = fine, `Not fine` = not_fine, `Fines unknown` = unknown,
            `Mean fines (%)` = mean_fines) |>
  kfmt()
```

## 3. Fraction distribution

The `grain_size_fraction` table holds the individual size fractions (corrected
percentages, with parsed `lo_um`/`hi_um` bounds) moved out of `measurement`.

```{r}
bind_rows(lapply(seq_len(nrow(gsrcs)), function(i)
  qry(gsrcs$stem[i], "SELECT COUNT(*) rows, COUNT(DISTINCT subsample_id) subsamples,
                      COUNT(DISTINCT symbol) codes FROM grain_size_fraction") |>
    mutate(Source = gsrcs$Source[i]))) |>
  transmute(Source, `Fraction rows` = rows, `Subsamples` = subsamples,
            `Distinct codes` = codes) |>
  kfmt()
```

::: {.callout-note}
`fines_lt63` and the fractions are derived, not deleted-from, quantities; the
grain-size correction (renormalising scaled curves, flagging unreliable values)
was applied in the slim stage and is documented on the
[slim site](https://seafood-hazards.github.io/multised-slim/).
:::
