---
title: "Harmonisation"
subtitle: "Uniform symbols, units and depths across the five sources"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Step 1 puts every source into one format, using the ICES-DOME vocabulary. Values
are not changed: the original value and unit are kept, alongside a standardised
value in a common unit.

```{r}
#| label: setup
library(DBI)
library(RSQLite)
library(tibble)
library(dplyr)
library(knitr)

db_dir <- "../data/db"
srcs <- tibble(
  Source = c("Mareano", "Vannmiljø", "ICES-DOME", "MUDAB", "4Demon"),
  stem   = c("mareano", "vannmiljo", "ices_dome", "mudab", "4demon"))

qry <- function(stem, sql) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(stem, "_clean.sqlite")))
  on.exit(dbDisconnect(con)); as_tibble(dbGetQuery(con, sql))
}
kfmt <- function(df, ...) kable(df, format.args = list(big.mark = ","), ...)
```

## 1. Symbols

Chemistry symbols are unified to the ICES names: the seven targets, the Fe/Al
normalisers, and organic carbon `CORG` (Mareano/Vannmiljø `TOC` maps to `CORG`;
Vannmiljø's `TOC63`, carbon on the `<63 µm` fraction, is kept separate). Case is
folded (`Fe` becomes `FE`). The chemistry symbols now present per source:

```{r}
bind_rows(lapply(seq_len(nrow(srcs)), function(i)
  qry(srcs$stem[i], "SELECT DISTINCT symbol FROM measurement ORDER BY symbol") |>
    summarise(Symbols = paste(symbol, collapse = ", ")) |>
    mutate(Source = srcs$Source[i]))) |>
  select(Source, Symbols) |>
  kable()
```

## 2. Units

The original `unit` is retained (renamed to the ICES form: micro sign to `u`, and
Vannmiljø's `dw` / `C` suffixes stripped), and every measurement also carries a
standardised `value_std` in **mg/kg** (`unit_std`), converted via its mass basis.
Original chemistry units still present:

```{r}
bind_rows(lapply(seq_len(nrow(srcs)), function(i)
  qry(srcs$stem[i], "SELECT DISTINCT unit FROM measurement WHERE unit IS NOT NULL ORDER BY unit") |>
    summarise(`Original units` = paste(unit, collapse = ", ")) |>
    mutate(Source = srcs$Source[i]))) |>
  select(Source, `Original units`) |>
  kable()
```

## 3. Depth

Sediment depth intervals are all converted to **centimetres**. The sources stored
depth differently: ICES-DOME in metres (multiplied by 100), the others already in
cm. Vannmiljø's corrupt depths (beyond a plausible core length, or inverted) were
set to unknown.

```{r}
bind_rows(lapply(seq_len(nrow(srcs)), function(i)
  qry(srcs$stem[i], "SELECT ROUND(MIN(depth_from),1) dmin, ROUND(MAX(depth_to),1) dmax,
                     SUM(depth_to IS NULL) n_unknown FROM subsample") |>
    mutate(Source = srcs$Source[i]))) |>
  transmute(Source, `Min depth (cm)` = dmin, `Max depth (cm)` = dmax,
            `Depth unknown` = n_unknown) |>
  kfmt()
```

## 4. Gear and detection limits

`sampling_tool` and `method` are carried in the ICES form where the source already
uses ICES-style codes (Mareano, ICES-DOME, MUDAB, 4Demon). Vannmiljø records ISO
sampling standards that do not map to ICES gear, so they are left as-is. Detection
and quantification limits (`lod` / `loq`) are kept on the `method` table with their
reporting unit (`limit_unit`); Mareano's single `lld` becomes `lod`, and 4Demon
carries no numeric limits.

```{r}
limit_avail <- function(stem) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(stem, "_clean.sqlite")))
  on.exit(dbDisconnect(con))
  flds <- dbListFields(con, "method")   # Mareano has no loq; 4Demon has neither
  n1 <- function(col) if (col %in% flds)
    as.integer(dbGetQuery(con, paste0("SELECT SUM(", col, " IS NOT NULL) FROM method"))[[1]]) else 0L
  tibble(Methods = as.integer(dbGetQuery(con, "SELECT COUNT(*) FROM method")[[1]]),
         `With LOD` = n1("lod"), `With LOQ` = n1("loq"))
}
bind_rows(lapply(seq_len(nrow(srcs)), function(i)
  limit_avail(srcs$stem[i]) |> mutate(Source = srcs$Source[i]))) |>
  select(Source, Methods, `With LOD`, `With LOQ`) |>
  kfmt()
```
