---
title: "multised (clean)"
subtitle: "The QC-passed, harmonised marine-sediment database from five sources"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

**multised** assembles marine-sediment trace-element data from five sources,
harmonises it into a common schema, and quality-controls it. This site documents
the **clean** stage: the QC-passed databases built from the slim stage by applying
the accumulated flags and standardising to a uniform format. It is the counterpart
of the [slim site](https://seafood-hazards.github.io/multised-slim/), which
documents the flagging that this stage acts on.

Each source becomes one `\<source\>_clean.sqlite`, still per-source (a cross-source
merge is a later step). The clean stage runs three steps:

```{r}
library(DBI)
library(RSQLite)
library(tibble)
library(dplyr)
library(knitr)

db_dir <- "../data/db"
srcs <- tibble(
  Source = c("Mareano", "Vannmiljø", "ICES-DOME", "MUDAB", "4Demon"),
  stem   = c("mareano", "vannmiljo", "ices_dome", "mudab", "4demon"))

scal <- function(stem, gen, sql) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(stem, "_", gen, ".sqlite")))
  on.exit(dbDisconnect(con)); as.integer(dbGetQuery(con, sql)[[1]])
}
kfmt <- function(df, ...) kable(df, format.args = list(big.mark = ","), ...)

tribble(
  ~Step,           ~Does,
  "1. Harmonise",  "Uniform symbols, units, depths (cm), tools, LOD/LOQ (ICES-DOME names).",
  "2. Clean",      "Remove failed-QC rows; keep one of each duplicate; average technical replicates.",
  "3. Annotate",   "Keep the labels; split grain-size into its own tables."
) |> kable()
```

## Data sources

```{r}
tribble(
  ~Source,     ~Origin,         ~`In clean stage`,
  "Mareano",   "Norway / IMR",  "steps 1-3",
  "Vannmiljø", "Norway",        "steps 1-3",
  "ICES-DOME", "International",  "steps 1-3 (naming reference)",
  "MUDAB",     "International",  "steps 1-3",
  "4Demon",    "International",  "steps 1-2 (no grain-size)"
) |> kable()
```

## What the cleaning removed

Chemistry measurements before (slim) and after (clean) the cleaning step. The
drop is the combined effect of removing failed-QC rows (out of range, invalid,
below limit, wet weight, source-specific QC) and collapsing duplicates and
technical replicates into one value each.

```{r}
chem_slim <- "SELECT COUNT(*) FROM measurement m JOIN element e ON m.symbol=e.symbol
              WHERE e.category IN ('target','reference','organic')"
srcs |>
  rowwise() |>
  mutate(`Chemistry (slim)`  = scal(stem, "slim",  chem_slim),
         `Chemistry (clean)` = scal(stem, "clean", "SELECT COUNT(*) FROM measurement")) |>
  ungroup() |>
  mutate(`% retained` = round(100 * `Chemistry (clean)` / `Chemistry (slim)`)) |>
  select(-stem) |>
  kfmt()
```

## Clean database tables

Every clean database has the seven shared tables, and the grain-size sources add
two more:

```{r}
tribble(
  ~Table,                 ~Contents,
  "element",              "Symbol, name, measurand category (ICES names).",
  "dataset / site",       "Provenance and location.",
  "event",                "Sampling occasion: date, year, gear.",
  "subsample",            "Depth interval (cm) + supporting-data flags.",
  "measurement",          "Chemistry only: value + standardised value_std (mg/kg), value_sd / n_rep, value_uncrt.",
  "method",               "Analysis + LOD/LOQ with their unit.",
  "grain_size",           "Per subsample: sieved/bulk, fines <63 um, is_fine.",
  "grain_size_fraction",  "The grain-size distribution moved out of measurement."
) |> kable()
```

::: {.callout-note}
The clean databases are gitignored working files; the pages here read them live at
render time, so all counts reflect the current build.
:::
