---
title: "Class Summary"
subtitle: Measurand classes and their value distributions across the five slim databases
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Before any quality flag is applied, each element is assigned a **measurand
class** — a single `category` column on the `element` table that every later
step reads instead of maintaining its own list of symbols. Four classes cover
the data:

- **Target** — the 7 trace-element targets: Co, Cu, I, Mn, Mo, Se, Zn.
- **Reference** — the Fe and Al normalisers.
- **Organic** — organic carbon (CORG / TOC / TOC63).
- **Composition** — grain-size parameters (the remainder).

This page summarises how the measurements split across those classes in each
source, the distribution of their standardised values, and which supporting data
(normalisers, organic carbon, grain-size composition) is available per sample.

```{r}
#| label: setup
library(DBI)
library(RSQLite)
library(tibble)
library(dplyr)
library(tidyr)
library(knitr)

db_dir <- "../data/db"

srcs <- tibble(
  Source = c("Mareano", "Vannmiljø", "ICES-DOME", "MUDAB", "4Demon"),
  file   = c("mareano_slim", "vannmiljo_slim", "ices_dome_slim",
             "mudab_slim", "4demon_slim")
)

scal <- function(file, sql) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(file, ".sqlite")))
  on.exit(dbDisconnect(con))
  as.integer(dbGetQuery(con, sql)[[1]])
}

qry <- function(file, sql) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(file, ".sqlite")))
  on.exit(dbDisconnect(con))
  as_tibble(dbGetQuery(con, sql))
}

add_total <- function(df) {
  tot <- df |> summarise(across(where(is.numeric), sum)) |> mutate(Source = "Total")
  bind_rows(df, tot)
}

kfmt <- function(df, ...) kable(df, format.args = list(big.mark = ","), ...)

# Format a summary statistic to 3 significant figures, comma-grouped, never
# scientific — so a table can mix 0.05 and 90,000,000,000 legibly. Sub-nanoscale
# magnitudes are floating-point noise (e.g. a grain-size % of -1.3e-14) and snap
# to 0.
fmt_val <- function(x) {
  x <- ifelse(abs(x) < 1e-9, 0, x)
  prettyNum(signif(x, 3), big.mark = ",", scientific = FALSE)
}

# Fixed class order for display.
cls_levels <- c("target", "reference", "organic", "composition")
cls_label  <- c(target = "Target", reference = "Reference",
                organic = "Organic", composition = "Composition")
cls_named  <- unname(cls_label[cls_levels])   # "Target","Reference",...
```

## 1. Measurements by class

Every measurement inherits its class from its element. The table shows how many
measurements fall in each class per source.

```{r}
meas_by_class <- bind_rows(lapply(seq_len(nrow(srcs)), function(i) {
  qry(srcs$file[i],
      "SELECT e.category AS category, COUNT(*) AS n
       FROM measurement m JOIN element e ON e.symbol = m.symbol
       GROUP BY e.category") |>
    mutate(Source = srcs$Source[i])
}))

meas_by_class |>
  mutate(category = factor(category, levels = cls_levels)) |>
  pivot_wider(names_from = category, values_from = n, values_fill = 0) |>
  mutate(Source = factor(Source, levels = srcs$Source)) |>
  arrange(Source) |>
  mutate(Source = as.character(Source)) |>
  rename_with(~ cls_label[.x], any_of(cls_levels)) |>
  relocate(Source, all_of(cls_named)) |>
  add_total() |>
  kfmt()
```

The number of *distinct elements* backing those measurements differs too — one
target set is largely shared, while composition spans many grain-size codes that
vary by source.

```{r}
el_by_class <- bind_rows(lapply(seq_len(nrow(srcs)), function(i) {
  qry(srcs$file[i],
      "SELECT category, COUNT(*) AS n FROM element GROUP BY category") |>
    mutate(Source = srcs$Source[i])
}))

el_by_class |>
  mutate(category = factor(category, levels = cls_levels)) |>
  pivot_wider(names_from = category, values_from = n, values_fill = 0) |>
  mutate(Source = factor(Source, levels = srcs$Source)) |>
  arrange(Source) |>
  mutate(Source = as.character(Source)) |>
  rename_with(~ cls_label[.x], any_of(cls_levels)) |>
  relocate(Source, all_of(cls_named)) |>
  kfmt()
```

## 2. Standardised-value distribution by class

Each measurement's standardised value `value_std` (chemistry in **mg/kg**,
composition in **%** — see *Unit Conversion & Range Check*) is summarised per
class and source. Chemistry classes are therefore comparable across sources on
one scale; composition is a percentage. Below-limit and unconvertible rows carry
no `value_std` and are excluded.

```{r}
vals <- bind_rows(lapply(seq_len(nrow(srcs)), function(i) {
  qry(srcs$file[i],
      "SELECT e.category AS category, m.value_std AS value_std
       FROM measurement m JOIN element e ON e.symbol = m.symbol
       WHERE m.value_std IS NOT NULL") |>
    mutate(Source = srcs$Source[i])
}))

vals |>
  mutate(category = factor(category, levels = cls_levels)) |>
  group_by(Source, Class = category) |>
  summarise(
    Unit     = if (first(Class) == "composition") "%" else "mg/kg",
    N        = n(),
    Min      = fmt_val(min(value_std)),
    Q1       = fmt_val(quantile(value_std, 0.25)),
    Median   = fmt_val(median(value_std)),
    Q3       = fmt_val(quantile(value_std, 0.75)),
    Max      = fmt_val(max(value_std)),
    .groups = "drop"
  ) |>
  mutate(Source = factor(Source, levels = srcs$Source),
         Class  = factor(cls_label[as.character(Class)], levels = cls_named)) |>
  arrange(Source, Class) |>
  mutate(Source = as.character(Source), Class = as.character(Class)) |>
  kfmt()
```

## 3. Supporting-data availability

Which data needed for later normalisation and characterisation is present in each
depth interval (`subsample`): an **Fe** or **Al** normaliser, **organic carbon**,
or **grain-size composition**. The counts are subsamples for which each is
available. These existence flags read the element class above (organic and
composition) and the Fe / Al symbols directly. (4Demon carries no organic-carbon
or composition parameters, so those are 0.)

```{r}
srcs |>
  rowwise() |>
  mutate(
    Subsamples  = scal(file, "SELECT COUNT(*) FROM subsample"),
    Fe          = scal(file, "SELECT COUNT(*) FROM subsample WHERE fe_exist = 1"),
    Al          = scal(file, "SELECT COUNT(*) FROM subsample WHERE al_exist = 1"),
    `Organic C` = scal(file, "SELECT COUNT(*) FROM subsample WHERE org_exist = 1"),
    Composition = scal(file, "SELECT COUNT(*) FROM subsample WHERE comp_exist = 1")
  ) |>
  ungroup() |>
  select(-file) |>
  add_total() |>
  kfmt()
```

::: {.callout-note}
The `category` column is descriptive, not a quality flag: no rows are marked or
removed here. It is the shared classification the flagging steps build on.
:::
