---
title: "Cleaning summary"
subtitle: "Chemistry measurements before (slim) and after (clean) cleaning"
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The cleaning step removes failed-QC rows (out of range, invalid, below limit, wet
weight, source-specific QC) and collapses duplicates and technical replicates into
one value each. This page counts the chemistry (target + reference + organic)
measurements before and after, overall and per element.

```{r}
#| label: setup
library(DBI)
library(RSQLite)
library(tibble)
library(dplyr)
library(tidyr)
library(knitr)

db_dir <- "../data/db"
srcs <- tibble(
  Source = c("Mareano", "Vannmiljø", "ICES-DOME", "MUDAB", "4Demon"),
  stem   = c("mareano", "vannmiljo", "ices_dome", "mudab", "4demon"))

qry <- function(stem, gen, sql) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(stem, "_", gen, ".sqlite")))
  on.exit(dbDisconnect(con)); as_tibble(dbGetQuery(con, sql))
}
scal <- function(stem, gen, sql) as.integer(qry(stem, gen, sql)[[1]])
kfmt <- function(df, ...) kable(df, format.args = list(big.mark = ","), ...)
```

## 1. Overall

Total chemistry measurements per source. The drop is the combined effect of the
removals and of collapsing duplicates and technical replicates.

```{r}
chem_slim <- "SELECT COUNT(*) FROM measurement m JOIN element e ON m.symbol=e.symbol
              WHERE e.category IN ('target','reference','organic')"
srcs |>
  rowwise() |>
  mutate(`Chemistry (slim)`  = scal(stem, "slim",  chem_slim),
         `Chemistry (clean)` = scal(stem, "clean", "SELECT COUNT(*) FROM measurement")) |>
  ungroup() |>
  mutate(`% retained` = round(100 * `Chemistry (clean)` / `Chemistry (slim)`)) |>
  select(-stem) |>
  kfmt()
```

## 2. Per element

The same counts broken down by element. The clean stage renames organic carbon
`TOC` to the ICES `CORG`, so the slim `TOC` rows are matched to clean `CORG` here
(the finer `TOC63` fraction keeps its own symbol).

```{r}
#| results: asis
slim_by_el <- "SELECT m.symbol AS Element, e.category AS Category, COUNT(*) AS n
               FROM measurement m JOIN element e ON m.symbol = e.symbol
               WHERE e.category IN ('target','reference','organic')
               GROUP BY m.symbol, e.category"
clean_by_el <- "SELECT symbol AS Element, COUNT(*) AS n FROM measurement GROUP BY symbol"

cat("::: {.panel-tabset}\n\n")
for (i in seq_len(nrow(srcs))) {
  s <- qry(srcs$stem[i], "slim", slim_by_el) |>
    mutate(Element = ifelse(toupper(Element) == "TOC", "CORG", toupper(Element))) |>
    group_by(Element, Category) |>
    summarise(n_slim = sum(n), .groups = "drop")
  cl <- qry(srcs$stem[i], "clean", clean_by_el) |>
    transmute(Element = toupper(Element), n_clean = n)
  tbl <- s |>
    left_join(cl, by = "Element") |>
    transmute(
      Element, Category,
      `Before (slim)` = n_slim,
      `After (clean)` = coalesce(n_clean, 0L),
      `% retained`    = round(100 * coalesce(n_clean, 0L) / n_slim)) |>
    arrange(Category, Element)

  cat("### ", srcs$Source[i], "\n\n")
  cat(kfmt(tbl, format = "pipe"), sep = "\n")
  cat("\n\n")
}
cat(":::\n\n")
```

::: {.callout-note}
The clean databases are gitignored working files; this page reads them live at
render time, so the counts reflect the current build.
:::
