---
title: "Common"
subtitle: Common flagging columns shared by all five slim databases
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

After each source is reshaped into the common slim schema, a series of **flagging
steps** annotates the data for quality control and review. Every step *adds*
columns; it never changes or deletes a value. Suspicious or redundant records
are marked so they can be inspected; the actual removal of unnecessary rows is a
later **Data Cleaning** step, performed once the grain-size analysis is complete.

This page covers the **common** flags: the columns that all five sources end up
with, applied the same way everywhere. The source-native flags that only some
sources report (e.g. detection flags and outlier flags) are summarised on the
companion *Source specific* page. Each section below gives a short description
followed by a cross-source count table, computed directly from the current slim
databases.

```{r}
#| label: setup
library(DBI)
library(RSQLite)
library(tibble)
library(dplyr)
library(knitr)

# Slim databases (internal working copies). Path is relative to this site's
# project root; the resulting counts are baked into the rendered HTML.
db_dir <- "../data/db"

srcs <- tibble(
  Source = c("Mareano", "Vannmiljø", "ICES-DOME", "MUDAB", "4Demon"),
  file   = c("mareano_slim", "vannmiljo_slim", "ices_dome_slim",
             "mudab_slim", "4demon_slim")
)

# Run a scalar-returning query against one source's slim database.
scal <- function(file, sql) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(file, ".sqlite")))
  on.exit(dbDisconnect(con))
  as.integer(dbGetQuery(con, sql)[[1]])
}

# Append a "Total" row summing every numeric column.
add_total <- function(df) {
  tot <- df |>
    summarise(across(where(is.numeric), sum)) |>
    mutate(Source = "Total")
  bind_rows(df, tot)
}

kfmt <- function(df, ...) kable(df, format.args = list(big.mark = ","), ...)
pct  <- function(n, d) ifelse(d == 0, "-", sprintf("%.1f%%", 100 * n / d))
```

## 1. Invalid values

Adds an `area_flag` to `site` and an `invalid_flag` to `measurement`. A site is
flagged **outside Europe** when its coordinates fall outside the European seas of
interest. A measurement is flagged **negative** when its value is below zero, or
**over-range** when it exceeds the physical ceiling of 100 % of the sample mass
for its unit (e.g. a `mg/kg` value above 1,000,000). These are candidates for
review, not automatic deletions.

```{r}
srcs |>
  rowwise() |>
  mutate(
    Sites            = scal(file, "SELECT COUNT(*) FROM site"),
    `Outside Europe` = scal(file, "SELECT COUNT(*) FROM site WHERE area_flag = 'outside_europe'"),
    Measurements     = scal(file, "SELECT COUNT(*) FROM measurement"),
    Negative         = scal(file, "SELECT COUNT(*) FROM measurement WHERE invalid_flag = 'negative'"),
    `Over-range`     = scal(file, "SELECT COUNT(*) FROM measurement WHERE invalid_flag = 'over_range'")
  ) |>
  ungroup() |>
  select(-file) |>
  add_total() |>
  kfmt()
```

## 2. Duplicates & technical replicates

Adds a `dup_flag` to `measurement`. Records sharing the same location, date,
depth, element and unit are grouped: if the value is also identical the row is a
**duplicate**; if the value differs it is a **technical replicate**. Both are
kept for now and flagged for the clean stage to resolve.

```{r}
srcs |>
  rowwise() |>
  mutate(
    Measurements           = scal(file, "SELECT COUNT(*) FROM measurement"),
    Duplicate              = scal(file, "SELECT COUNT(*) FROM measurement WHERE dup_flag = 'duplicate'"),
    `Technical replicate`  = scal(file, "SELECT COUNT(*) FROM measurement WHERE dup_flag = 'technical_replicate'")
  ) |>
  ungroup() |>
  select(-file) |>
  add_total() |>
  kfmt()
```

## 3. Supporting-data availability

Adds four existence flags to `subsample`, recording whether the data needed for
later normalisation and characterisation is present in each depth interval: an
**Fe** or **Al** normaliser, **organic carbon** (CORG / TOC), or **grain-size
composition**. The counts below are subsamples for which each is available.
(4Demon carries no organic-carbon or composition parameters, so those are 0.)

```{r}
srcs |>
  rowwise() |>
  mutate(
    Subsamples      = scal(file, "SELECT COUNT(*) FROM subsample"),
    Fe              = scal(file, "SELECT COUNT(*) FROM subsample WHERE fe_exist = 1"),
    Al              = scal(file, "SELECT COUNT(*) FROM subsample WHERE al_exist = 1"),
    `Organic C`     = scal(file, "SELECT COUNT(*) FROM subsample WHERE org_exist = 1"),
    Composition     = scal(file, "SELECT COUNT(*) FROM subsample WHERE comp_exist = 1")
  ) |>
  ungroup() |>
  select(-file) |>
  add_total() |>
  kfmt()
```

## 4. Multi-layer samplings

Adds `n_layers` and `multi_flag` to `event`. An event with more than one
subsample is a **multi-layer** sampling (e.g. a sliced core), while a single
depth interval is a **single grab**. This is derived from the data, since the
same gear yields both. The last column is the largest number of layers seen in
any single event.

```{r}
tbl6 <- srcs |>
  rowwise() |>
  mutate(
    Events        = scal(file, "SELECT COUNT(*) FROM event"),
    `Multi-layer` = scal(file, "SELECT COUNT(*) FROM event WHERE multi_flag = 1"),
    `Single grab` = scal(file, "SELECT COUNT(*) FROM event WHERE multi_flag = 0"),
    `Max layers`  = scal(file, "SELECT MAX(n_layers) FROM event")
  ) |>
  ungroup() |>
  select(-file)

# Total row: sum the counts, take the overall max for the layer depth.
bind_rows(
  tbl6,
  tibble(Source = "Total",
         Events = sum(tbl6$Events),
         `Multi-layer` = sum(tbl6$`Multi-layer`),
         `Single grab` = sum(tbl6$`Single grab`),
         `Max layers`  = max(tbl6$`Max layers`))
) |>
  kfmt()
```

## 5. Below detection / quantification limit

Adds a `below_loq` flag to `measurement`, folding each source's own detection
flag (Mareano `below_lld`, Vannmiljø `operator`, the ICES `qflag`, 4Demon
`limit_flag`) into one common marker. A value of 1 means the measurement is an
uncertain "less-than" reading below the limit of detection or quantification,
the main set of rows to remove in the clean stage.

```{r}
srcs |>
  rowwise() |>
  mutate(
    Measurements = scal(file, "SELECT COUNT(*) FROM measurement"),
    `Below LOQ`  = scal(file, "SELECT COUNT(*) FROM measurement WHERE below_loq = 1")
  ) |>
  ungroup() |>
  mutate(Share = pct(`Below LOQ`, Measurements)) |>
  select(-file) |>
  kfmt()
```

::: {.callout-note}
No rows have been deleted. These flags will drive the removal of uncertain,
duplicated, and out-of-scope records when the **clean** databases are built.
:::
