---
title: "Source Specific"
subtitle: Native flags that only some sources carry
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The *Common Flags* are applied the same way to all five sources. Some sources
also carry **native columns of their own** that the common steps do not cover.
Where those add information worth keeping, they are folded into a per-source
`src_flag` on `measurement` (NULL = pass), so the clean stage can review or remove
them. This is a source-by-source step (13 onward in the pipeline); so far
**Vannmiljø** and **ICES-DOME** have one.

```{r}
#| label: setup
library(DBI)
library(RSQLite)
library(tibble)
library(dplyr)
library(knitr)

db_dir <- "../data/db"

qry <- function(file, sql) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(file, ".sqlite")))
  on.exit(dbDisconnect(con))
  as_tibble(dbGetQuery(con, sql))
}

kfmt <- function(df, ...) kable(df, format.args = list(big.mark = ","), ...)
```

## Vannmiljø

Vannmiljø records two native columns on each measurement that the other sources
do not: `operator` (a relational sign on the value) and `filtered`. The
below-detection meaning of `operator` (`<` and `ND`) is already captured by the
common `below_loq` flag, so it is not repeated here. What remains is folded into
`src_flag`:

```{r}
tribble(
  ~`src_flag`,     ~`Source column`,  ~Meaning,
  "above_range", "operator = '>'",  "Right-censored 'greater-than' reading: the value is only a lower bound.",
  "filtered",    "filtered = 1",    "Filtered-water sample rather than bulk sediment."
) |> kable()
```

Neither is scoped to a measurand class. In particular `above_range` also applies
to grain-size fractions reported as "greater than" a value, not just chemistry.
The counts below are the current Vannmiljø slim database.

```{r}
qry("vannmiljo_slim",
    "SELECT COALESCE(src_flag, '(pass)') AS `src_flag`, COUNT(*) AS Rows
     FROM measurement GROUP BY src_flag ORDER BY Rows DESC") |>
  kfmt()
```

For context, the full `operator` breakdown (its `<` / `ND` values feed
`below_loq`, `>` feeds `src_flag`, `=` is an ordinary quantified value):

```{r}
qry("vannmiljo_slim",
    "SELECT operator AS Operator, COUNT(*) AS Rows
     FROM measurement GROUP BY operator ORDER BY Rows DESC") |>
  kfmt()
```

## ICES-DOME

ICES-DOME's detection/quantification flag (`qflag`) and dry/wet basis (`basis`)
are already handled by the common `below_loq` and `weight_basis` steps. Its
`vflag`, the originator's value-quality flag, is folded into `src_flag` (meanings
from the ICES `code_lookup` table):

```{r}
tribble(
  ~`src_flag`,    ~`Source column`, ~Meaning,
  "suspect",    "vflag = 'S'",  "Suspect value: considered suspect by the originator on the basis of quality control or instrument performance.",
  "calculated", "vflag = 'C'",  "Calculated value: a derived value rather than a direct measurement."
) |> kable()
```

`vflag = 'A'` (acceptable value) and rows with no `vflag` pass. The counts below
are the current ICES-DOME slim database.

```{r}
qry("ices_dome_slim",
    "SELECT COALESCE(src_flag, '(pass)') AS `src_flag`, COUNT(*) AS Rows
     FROM measurement GROUP BY src_flag ORDER BY Rows DESC") |>
  kfmt()
```

ICES-DOME's remaining native columns are **not** folded in: `dcflag` holds ICES
DATSU screening / conversion codes (mostly benign unit conversions), and
`metcu` / `uncrt` / `matrix` are uncertainty and sample-fraction metadata rather
than quality flags.

## Other sources

MUDAB and 4Demon carry their own native columns too (e.g. 4Demon's
`range_check_flag` / `outlier_extreme_flag` / `outlier_stdev_flag`). These are
retained in the slim `measurement` tables but are **not yet folded into a
`src_flag`**; they will be added here as their handling is decided.

::: {.callout-note}
No rows have been deleted. `src_flag` marks source-native records for review; the
actual removal happens later, when the **clean** databases are built.
:::
