---
title: "Source Specific"
subtitle: Native flags that only some sources carry
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The *Common Flags* are applied the same way to all five sources. Some sources
also carry **native columns of their own** that the common steps do not cover.
Where those add information worth keeping, they are folded into a per-source
`src_flag` on `measurement` (NULL = pass), so the clean stage can review or remove
them. This is a source-by-source step (13 onward in the pipeline); so far only
**Vannmiljø** has one.

```{r}
#| label: setup
library(DBI)
library(RSQLite)
library(tibble)
library(dplyr)
library(knitr)

db_dir <- "../data/db"

qry <- function(file, sql) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(file, ".sqlite")))
  on.exit(dbDisconnect(con))
  as_tibble(dbGetQuery(con, sql))
}

kfmt <- function(df, ...) kable(df, format.args = list(big.mark = ","), ...)
```

## Vannmiljø

Vannmiljø records two native columns on each measurement that the other sources
do not: `operator` (a relational sign on the value) and `filtered`. The
below-detection meaning of `operator` (`<` and `ND`) is already captured by the
common `below_loq` flag, so it is not repeated here. What remains is folded into
`src_flag`:

```{r}
tribble(
  ~`src_flag`,     ~`Source column`,  ~Meaning,
  "above_range", "operator = '>'",  "Right-censored 'greater-than' reading: the value is only a lower bound.",
  "filtered",    "filtered = 1",    "Filtered-water sample rather than bulk sediment."
) |> kable()
```

Neither is scoped to a measurand class. In particular `above_range` also applies
to grain-size fractions reported as "greater than" a value, not just chemistry.
The counts below are the current Vannmiljø slim database.

```{r}
qry("vannmiljo_slim",
    "SELECT COALESCE(src_flag, '(pass)') AS `src_flag`, COUNT(*) AS Rows
     FROM measurement GROUP BY src_flag ORDER BY Rows DESC") |>
  kfmt()
```

For context, the full `operator` breakdown (its `<` / `ND` values feed
`below_loq`, `>` feeds `src_flag`, `=` is an ordinary quantified value):

```{r}
qry("vannmiljo_slim",
    "SELECT operator AS Operator, COUNT(*) AS Rows
     FROM measurement GROUP BY operator ORDER BY Rows DESC") |>
  kfmt()
```

## Other sources

The international sources carry their own native columns (e.g. ICES-DOME's
`vflag` / `dcflag` / `metcu` / `uncrt`, and 4Demon's `range_check_flag` /
`outlier_extreme_flag` / `outlier_stdev_flag`). These are retained in the slim
`measurement` tables but are **not yet folded into a `src_flag`**; they will be
added here as their handling is decided.

::: {.callout-note}
No rows have been deleted. `src_flag` marks source-native records for review; the
actual removal happens later, when the **clean** databases are built.
:::
