---
title: "Unit Conversion & Range Check"
subtitle: Standardising values to a common unit, then flagging implausible concentrations
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The five sources report the same measurements in different units: `mg/kg`,
`µg/kg`, `%`, `g/kg`, and so on. Before implausible values can be flagged
consistently, every value is first **converted to a common unit**, and the
standardised chemistry values are then compared against a **plausible range** per
element. Both are additive columns on `measurement`; nothing is deleted here.

```{r}
#| label: setup
library(DBI)
library(RSQLite)
library(tibble)
library(dplyr)
library(tidyr)
library(knitr)

db_dir <- "../data/db"

srcs <- tibble(
  Source = c("Mareano", "Vannmiljø", "ICES-DOME", "MUDAB", "4Demon"),
  file   = c("mareano_slim", "vannmiljo_slim", "ices_dome_slim",
             "mudab_slim", "4demon_slim")
)

scal <- function(file, sql) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(file, ".sqlite")))
  on.exit(dbDisconnect(con))
  as.integer(dbGetQuery(con, sql)[[1]])
}

qry <- function(file, sql) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(file, ".sqlite")))
  on.exit(dbDisconnect(con))
  as_tibble(dbGetQuery(con, sql))
}

add_total <- function(df) {
  tot <- df |> summarise(across(where(is.numeric), sum)) |> mutate(Source = "Total")
  bind_rows(df, tot)
}

kfmt <- function(df, ...) kable(df, format.args = list(big.mark = ","), ...)
```

## 1. Unit conversion

Each measurement gets a standardised value `value_std` and its unit `unit_std`,
assigned by **measurand** rather than by the raw unit:

- **Chemistry**: the 7 target elements, the Fe/Al normalisers, and organic
  carbon (CORG / TOC / TOC63) → **mg/kg** dry weight.
- **Grain-size composition** → **%** (a mass/volume fraction). There is no
  length/µm grain-size in the data, so every grain-size value is a fraction and
  standardises to `%`.

The conversion goes through the mass fraction `value / denom`, where `denom` is
"100 % of the sample mass" expressed in the original unit, then scales to the
target unit's full scale (`1,000,000` for mg/kg, `100` for %).

Where a source supplies a QC-corrected value, that is the value standardised here:
4Demon's `corrected_value` (which its metadata recommends for analysis) carries
scale-error fixes and below-detection substitutions, while the raw `value` is kept
as provenance. This corrected a batch of 4Demon iron values reported about 10 times
too high.

```{r}
tribble(
  ~Measurand,                                ~`Standard unit`, ~`Example conversion`,
  "Targets (Co, Cu, I, Mn, Mo, Se, Zn)",     "mg/kg",          "9.8 % Cu → 98,000 mg/kg",
  "Normalisers (Fe, Al)",                    "mg/kg",          "3.31 % Al → 33,100 mg/kg",
  "Organic carbon (CORG / TOC / TOC63)",     "mg/kg",          "85 g/kg C → 85,000 mg/kg",
  "Grain-size composition",                  "%",              "180 g/kg → 18 %"
) |> kable()
```

The table below shows how each source's measurements split between the two
standardised units. Every unit present maps to a mass basis, so nothing is left
unconverted.

```{r}
srcs |>
  rowwise() |>
  mutate(
    Measurements        = scal(file, "SELECT COUNT(*) FROM measurement"),
    `Chemistry (mg/kg)` = scal(file, "SELECT COUNT(*) FROM measurement WHERE unit_std = 'mg/kg'"),
    `Grain-size (%)`    = scal(file, "SELECT COUNT(*) FROM measurement WHERE unit_std = '%'"),
    `Not converted`     = scal(file, "SELECT COUNT(*) FROM measurement WHERE unit_std IS NULL")
  ) |>
  ungroup() |>
  select(-file) |>
  add_total() |>
  kfmt()
```

## 2. Plausible range check

The standardised `value_std` (mg/kg) of every chemistry measurement is compared
against a per-element **plausible range**. A value below the minimum is flagged
`below_min`, above the maximum `above_max`; everything else, including grain-size
(unbounded) and below-LOQ readings (a limit, not a value), stays unflagged.

::: {.callout-warning}
These bounds are a **draft**, deliberately generous: they target *clearly
implausible* values (both impossibly high and impossibly low), not strict
geochemical limits, and are still to be reviewed with domain experts.
:::

```{r}
tribble(
  ~Element,                ~`Min (mg/kg)`, ~`Max (mg/kg)`,
  "Cobalt (Co)",           0.1,            300,
  "Copper (Cu)",           0.5,            10000,
  "Iodine (I)",            0.1,            1000,
  "Manganese (Mn)",        1,              50000,
  "Molybdenum (Mo)",       0.05,           500,
  "Selenium (Se)",         0.01,           100,
  "Zinc (Zn)",             1,              20000,
  "Iron (Fe)",             500,            250000,
  "Aluminium (Al)",        500,            200000,
  "Organic C (CORG/TOC/TOC63)", 100,       300000
) |> kfmt()
```

### Flagged values per source

```{r}
srcs |>
  rowwise() |>
  mutate(
    Measurements = scal(file, "SELECT COUNT(*) FROM measurement"),
    `Below min`  = scal(file, "SELECT COUNT(*) FROM measurement WHERE range_flag = 'below_min'"),
    `Above max`  = scal(file, "SELECT COUNT(*) FROM measurement WHERE range_flag = 'above_max'")
  ) |>
  ungroup() |>
  select(-file) |>
  add_total() |>
  kfmt()
```

### Flagged values per element (all sources)

```{r}
elname <- tribble(
  ~sym,     ~Element,
  "CO", "Cobalt", "CU", "Copper", "I", "Iodine", "MN", "Manganese",
  "MO", "Molybdenum", "SE", "Selenium", "ZN", "Zinc", "FE", "Iron",
  "AL", "Aluminium", "CORG", "Organic C", "TOC", "Organic C (TOC)",
  "TOC63", "Organic C (TOC63)"
)

tbl <- bind_rows(lapply(srcs$file, function(f)
  qry(f, "SELECT UPPER(symbol) sym, range_flag, COUNT(*) n
          FROM measurement WHERE range_flag IS NOT NULL
          GROUP BY UPPER(symbol), range_flag") |>
    transmute(sym = as.character(sym),
              range_flag = as.character(range_flag),
              n = as.integer(n)))) |>
  group_by(sym, range_flag) |>
  summarise(n = sum(n), .groups = "drop") |>
  pivot_wider(names_from = range_flag, values_from = n, values_fill = 0)

if (!"below_min" %in% names(tbl)) tbl$below_min <- 0L
if (!"above_max" %in% names(tbl)) tbl$above_max <- 0L

tbl |>
  left_join(elname, by = "sym") |>
  transmute(Element = coalesce(Element, sym),
            `Below min` = below_min,
            `Above max` = above_max) |>
  arrange(desc(`Below min` + `Above max`)) |>
  kfmt()
```

Flagged rows are review candidates: they carry the `range_flag` marker but keep
their values, and are removed only later, in the **Data Cleaning** step.
