---
title: "Fines (<63 µm)"
subtitle: The mud (clay + silt) fraction of each sample, derived from grain-size data
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

The fraction of a sediment finer than 63 µm (clay plus silt, the "mud" fraction)
is a key descriptor: it governs how much surface area is available to bind trace
elements, and it is the usual basis for comparing contaminant levels between
samples. This page derives a single per-sample value, `subsample.fines_lt63` (a
percentage), from each source's grain-size measurements, and records in
`fines_basis` how it was obtained.

Every source with grain-size contributes (all but 4Demon, which carries no
grain-size). The value is taken from the corrected standardised value
`value_std_corr` (grain size to %), so mixed reporting units (`%`, `g/kg`) are
handled uniformly and the grain-size curves repaired in the correction step (see
below) are used.

```{r}
#| label: setup
library(DBI)
library(RSQLite)
library(tibble)
library(dplyr)
library(tidyr)
library(knitr)

db_dir <- "../data/db"

srcs <- tibble(
  Source = c("Mareano", "Vannmiljø", "ICES-DOME", "MUDAB"),
  file   = c("mareano_slim", "vannmiljo_slim", "ices_dome_slim", "mudab_slim"))

scal <- function(file, sql) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(file, ".sqlite")))
  on.exit(dbDisconnect(con))
  as.integer(dbGetQuery(con, sql)[[1]])
}
qry <- function(file, sql) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(file, ".sqlite")))
  on.exit(dbDisconnect(con))
  as_tibble(dbGetQuery(con, sql))
}
kfmt <- function(df, ...) kable(df, format.args = list(big.mark = ","), ...)

# all fines values, tagged by source
vals <- bind_rows(lapply(seq_len(nrow(srcs)), function(i)
  qry(srcs$file[i], "SELECT fines_lt63 AS v, fines_basis AS basis
                     FROM subsample WHERE fines_lt63 IS NOT NULL") |>
    mutate(Source = srcs$Source[i]))) |>
  mutate(Source = factor(Source, levels = srcs$Source))
```

## 1. How the fines fraction is derived

Each source encodes grain-size differently, and the parameter names do not mean
the same thing across sources, so the meaning of every code was checked against
the pilot parameter definitions before use.

```{r}
tribble(
  ~Source,     ~`Code used`,                    ~`fines_lt63`,                                  ~`fines_basis`,
  "Mareano",   "Clay + Silt bins",              "Clay (<2 µm) + Silt (2 to 63 µm)",             "sum_bins",
  "Vannmiljø", "FINS, complement, or clay+silt", "FINS, else 100 minus GSMF_63, else GSMF2 + GSMF2_63", "fins / gsmf_63_complement / clay_silt_sum",
  "ICES-DOME", "GSMF63 on a bulk matrix",       "Cumulative <63 µm on SEDtot, else SED2000",    "gsmf63_sedtot / gsmf63_sed2000",
  "MUDAB",     "GSMF63 on a bulk matrix",       "Cumulative <63 µm on SEDtot, else SED2000",    "gsmf63_sedtot / gsmf63_sed2000"
) |> kable()
```

Two points that the derivation has to get right:

- **The names can be reversed.** In ICES-DOME and MUDAB, `GSMF63` is the mass
  fraction *finer* than 63 µm (silt and clay). In Vannmiljø the similar-looking
  `GSMF_63` is the *opposite*: "Particle fraction >63 µm" (sand and coarser). So
  Vannmiljø's fines come from `FINS`, and where only `GSMF_63` exists the fines
  are `100 minus GSMF_63`.
- **The fraction is relative to the matrix.** A cumulative `GSMF63` is the <63 µm
  share *of whatever fraction it was measured on* (see the Sample Fraction page).
  It is taken from the whole sample (`SEDtot`) first, and otherwise from the
  bulk-equivalent `SED2000` / `SED1000` (coarse sieving that only removes gravel).
  Values measured on an already-fine fraction (which would be near 100 %) are not
  used.

## 2. Correcting noisy grain-size curves

Much of the international grain-size data is noisy: many curves are internally
consistent (a monotone cumulative distribution) but scaled wrong, so a value that
should sit near 100 % is reported as thousands of percent, with every code in the
sample inflated by the same factor. A separate correction step handles this before
the fines are derived, writing a corrected value (`value_std_corr`) and a marker
(`gs_corr`) on each measurement:

- **ICES-DOME and MUDAB** renormalise each over-scaled, still-monotone curve so its
  coarsest cutoff (the total, near the `<2 mm` fraction) reads 100 %
  (`gs_corr = 'renorm'`). This recovers the affected samples for the fines below.
  Curves that stay implausible (not monotone) are marked `suspect` and excluded.
- **Vannmiljø** has only a handful of isolated bad values (and its codes mean
  ">n µm", so the curve renormalisation does not apply): these are flagged
  `suspect` for manual review, not rescaled.

```{r}
srcs |>
  filter(Source != "Mareano") |>   # Mareano is clean, has no gs_corr column
  rowwise() |>
  mutate(
    Renormed = scal(file, "SELECT COUNT(*) FROM measurement WHERE gs_corr = 'renorm'"),
    Suspect  = scal(file, "SELECT COUNT(*) FROM measurement WHERE gs_corr = 'suspect'")) |>
  ungroup() |>
  select(Source, Renormed, Suspect) |>
  kfmt()
```

Mareano needs no correction (its named-bin grain-size is already clean), so it has
no `gs_corr` column and its fines read the uncorrected value.

## 3. Coverage

```{r}
cov <- srcs |>
  rowwise() |>
  mutate(
    Subsamples   = scal(file, "SELECT COUNT(*) FROM subsample"),
    `With fines` = scal(file, "SELECT COUNT(*) FROM subsample WHERE fines_lt63 IS NOT NULL")) |>
  ungroup() |>
  mutate(`% covered` = round(100 * `With fines` / Subsamples)) |>
  select(-file)
kfmt(cov)
```

Coverage is partial by design: a subsample contributes only if it has a usable
`<63 µm` measurement. Implausible values (a cumulative mass fraction outside
0 to 100 %, about 1 % of the international grain-size rows) are left out rather
than corrected here.

The `fines_basis` breakdown shows which route each value came from:

```{r}
vals |>
  count(Source, Basis = basis, name = "Subsamples") |>
  arrange(Source, desc(Subsamples)) |>
  kfmt()
```

## 4. Distribution

Summary of the derived `fines_lt63` (%) per source:

```{r}
vals |>
  group_by(Source) |>
  summarise(
    N      = n(),
    Min    = round(min(v), 1),
    Q1     = round(quantile(v, 0.25), 1),
    Median = round(median(v), 1),
    Mean   = round(mean(v), 1),
    Q3     = round(quantile(v, 0.75), 1),
    Max    = round(max(v), 1),
    .groups = "drop") |>
  kfmt()
```

Spread across mud-content bands (share of each source's samples):

```{r}
bands <- c("0-10", "10-25", "25-50", "50-75", "75-90", "90-100")
vals |>
  mutate(Band = cut(v, breaks = c(-Inf, 10, 25, 50, 75, 90, Inf),
                    labels = bands, right = TRUE)) |>
  count(Source, Band) |>
  group_by(Source) |>
  mutate(pct = round(100 * n / sum(n))) |>
  ungroup() |>
  select(-n) |>
  pivot_wider(names_from = Band, values_from = pct, values_fill = 0) |>
  relocate(Source, all_of(bands)) |>
  kfmt()
```

::: {.callout-note}
`fines_lt63` is a reusable derived quantity, not a flag: nothing is deleted. For
Vannmiljø, samples with neither direct code fall back to the clay + silt sum
(`GSMF2` + `GSMF2_63`), the same construction as Mareano. The equivalent bin route
for ICES-DOME and MUDAB reaches only `<60 µm` (a different cutoff) and was left
out. Still to come: the gravel adjustment for `SED2000` based fines (the `<2 mm`
total is taken as the whole sample), and manual review of the `suspect` rows that
the correction could not repair.
:::
