---
title: "Sample Fraction (matrix)"
subtitle: What sediment fraction each measurement was made on, and how sieved and bulk samples are told apart
---

```{css, echo=FALSE}
th { white-space: nowrap; }
```

Sediment can be analysed **whole** (a bulk grab) or after **sieving** to a
grain-size fraction, and a trace-element value means something different
depending on which. Three sources record this on each measurement in a `matrix`
column: **ICES-DOME**, **MUDAB**, and **4Demon**. Mareano and Vannmiljø have no
`matrix`; for those the fraction has to come from the grain-size measurements
themselves (a later page).

This page reads the raw `matrix` values, harmonises them into a common
sieved/bulk class with a sieve cutoff in micrometres, and shows an important
subtlety: within one sample the chemistry and the grain-size are usually
measured on **different** fractions.

```{r}
#| label: setup
library(DBI)
library(RSQLite)
library(tibble)
library(dplyr)
library(tidyr)
library(knitr)

db_dir <- "../data/db"
chem_cats <- c("target", "reference", "organic")

kfmt <- function(df, ...) kable(df, format.args = list(big.mark = ","), ...)

# Sieve cutoff in micrometres, from either encoding:
#   ICES / MUDAB : matrix = "SED<n>"  -> n         ("SEDtot" has no cutoff)
#   4Demon       : fraction_range = "0-<n>" -> n   (matrix is only FS / US)
cut_um <- function(matrix, fraction_range) {
  fr  <- suppressWarnings(as.numeric(sub(".*-", "", fraction_range)))
  sed <- suppressWarnings(as.numeric(sub("^SED", "", matrix)))
  coalesce(fr, sed)
}

# Harmonised class. Coarse sieving (cutoff >= 1000 um, i.e. <2 mm / <1 mm) is the
# standard "remove gravel" prep and is treated as bulk-equivalent, together with
# the untreated total (SEDtot) and the unsieved 4Demon code (US). Finer cutoffs
# are the sieved fine fractions. PK_default and a missing matrix are unspecified.
frac_class <- function(matrix, cutoff) {
  case_when(
    is.na(matrix)               ~ "unspecified",
    matrix == "PK_default"      ~ "unspecified",
    matrix == "SEDtot"          ~ "bulk",
    matrix == "US"              ~ "bulk",
    !is.na(cutoff) & cutoff >= 1000 ~ "bulk",
    !is.na(cutoff)              ~ "sieved",
    TRUE                        ~ "unspecified"
  )
}

load_frac <- function(file) {
  con <- dbConnect(SQLite(), file.path(db_dir, paste0(file, ".sqlite")))
  on.exit(dbDisconnect(con))
  has_fr <- "fraction_range" %in% dbListFields(con, "measurement")
  fr_sel <- if (has_fr) "m.fraction_range" else "NULL AS fraction_range"
  sql <- sprintf(
    "SELECT m.subsample_id, m.matrix, %s, e.category
       FROM measurement m JOIN element e ON m.symbol = e.symbol", fr_sel)
  as_tibble(dbGetQuery(con, sql)) |>
    mutate(
      fraction_range = as.character(fraction_range),
      cutoff  = cut_um(matrix, fraction_range),
      class   = frac_class(matrix, cutoff),
      grp     = if_else(category %in% chem_cats, "Chemistry", "Grain-size"))
}

srcs <- c(`ICES-DOME` = "ices_dome_slim",
          MUDAB       = "mudab_slim",
          `4Demon`    = "4demon_slim")

frac <- bind_rows(lapply(names(srcs), function(nm)
  load_frac(srcs[[nm]]) |> mutate(Source = nm)))
```

## 1. What the matrix codes mean

**ICES-DOME and MUDAB** use ICES `MATRX` codes of the form `SED<mesh>`, "sediment,
the fraction finer than `<mesh>` micrometres" (`SEDtot` being the whole untreated
sample). Meanings are from each pilot's `code_lookup` table.

```{r}
tribble(
  ~matrix,     ~Meaning,                        ~`Cutoff (um)`,
  "SEDtot",    "Untreated total (whole sample)", "none (bulk)",
  "SED2000",   "Finer than 2000 um",             "2000",
  "SED1000",   "Finer than 1000 um",             "1000",
  "SED500",    "Finer than 500 um",              "500",
  "SED90",     "Finer than 90 um",               "90",
  "SED63",     "Finer than 63 um",               "63",
  "SED62",     "Finer than 62 um",               "62",
  "SED20",     "Finer than 20 um",               "20",
  "PK_default","Placeholder, fraction not stated","unspecified"
) |> kable()
```

**4Demon** records only `FS` (a sieved fraction) or `US` (unsieved), and keeps
the actual cutoff in a separate `fraction_range` column (micrometres), which is
carried through to the slim database.

```{r}
tribble(
  ~matrix, ~`fraction_range`, ~Meaning,
  "US",    "0-10000",         "Unsieved (whole sample)",
  "FS",    "0-2000",          "Sieved finer than 2000 um",
  "FS",    "0-500",           "Sieved finer than 500 um",
  "FS",    "0-63",            "Sieved finer than 63 um",
  "FS",    "0-37",            "Sieved finer than 37 um"
) |> kable()
```

There is no `SED64` code in any source: the fine cutoff of interest sits at 62
to 63 micrometres (`SED63`, `SED62`, `FS 0-63`).

## 2. Harmonised sieved / bulk class

Every measurement is placed in one of three classes from its `matrix` (and, for
4Demon, its `fraction_range`):

```{r}
tribble(
  ~Class,        ~Definition,                                                    ~`Codes`,
  "bulk",        "Whole sample, or coarse sieving at or above 1000 um",          "SEDtot, US, SED2000, SED1000, FS 0-2000",
  "sieved",      "Sieved to a fine fraction below 1000 um",                      "SED500, SED90, SED63, SED62, SED20, FS 0-500 / 0-63 / 0-37",
  "unspecified", "No fraction stated",                                           "PK_default, missing matrix"
) |> kable()
```

::: {.callout-note}
Coarse sieving to `<2 mm` or `<1 mm` (`SED2000`, `SED1000`, `FS 0-2000`) is the
routine step that removes gravel and shell before analysis, not a fine-fraction
separation, so it is counted as **bulk-equivalent** here rather than as a sieved
fraction.
:::

## 3. Fraction by source and measurement class

Because chemistry and grain-size are often run on different fractions (see
section 5), the split is shown separately for each. Counts are measurements.

```{r}
frac |>
  count(Source, grp, class) |>
  pivot_wider(names_from = class, values_from = n, values_fill = 0) |>
  relocate(Source, `Measurement class` = grp) |>
  arrange(Source, `Measurement class`) |>
  kfmt()
```

## 4. Chemistry sieve cutoffs

For the chemistry measurements, this is the actual sieve cutoff (micrometres) the
element value was measured on. The `<63 um` fine fraction is the one of primary
interest.

```{r}
frac |>
  filter(grp == "Chemistry", class == "sieved") |>
  count(Source, `Cutoff (um)` = cutoff) |>
  pivot_wider(names_from = `Cutoff (um)`, values_from = n, values_fill = 0) |>
  kfmt()
```

Chemistry measured as **bulk** (whole or coarse) for comparison:

```{r}
frac |>
  filter(grp == "Chemistry", class == "bulk") |>
  count(Source, name = "Bulk chemistry rows") |>
  kfmt()
```

## 5. The fraction differs between chemistry and grain-size

Within a single subsample the chemistry and the grain-size are usually **not** on
the same fraction: grain-size is almost always reported on the whole sample
(`SEDtot`), while chemistry is run on a sieved fine fraction. So there is no single
"matrix per sample"; the fraction has to be read per measurement, and separately
for chemistry versus grain-size.

```{r}
agree <- frac |>
  filter(Source != "4Demon") |>          # 4Demon has no grain-size
  group_by(Source, subsample_id, grp) |>
  summarise(mats = paste(sort(unique(matrix)), collapse = "|"), .groups = "drop") |>
  pivot_wider(names_from = grp, values_from = mats) |>
  filter(!is.na(Chemistry), !is.na(`Grain-size`))

agree |>
  group_by(Source) |>
  summarise(
    `Subsamples with both` = n(),
    `Same matrix`          = sum(Chemistry == `Grain-size`),
    `Different matrix`     = sum(Chemistry != `Grain-size`),
    .groups = "drop") |>
  mutate(`% different` = round(100 * `Different matrix` / `Subsamples with both`)) |>
  kfmt()
```

The most common mismatch is chemistry on a sieved fraction while the grain-size is
reported on the bulk sample:

```{r}
agree |>
  filter(Chemistry != `Grain-size`) |>
  count(Chemistry, `Grain-size`, name = "Subsamples") |>
  arrange(desc(Subsamples)) |>
  head(10) |>
  kfmt()
```

::: {.callout-note}
Nothing is deleted or recomputed on this page. It only reads and classifies the
recorded `matrix`. Recalculating noisy grain-size fractions, and deriving the
`<63 um` fraction for Mareano and Vannmiljø, come in later pages of this section.
:::
