---
title: "Analyses: merged"
description: >
  Thirteen cross-source analyses on the merged database, feeding the
  multised-merged site.
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

Once the five sources are one database, the questions change. The clean
analyses ask what each source holds; the merged analyses ask what the *combined*
record shows — where the concentrations cluster, which regions look alike, how a
profile behaves down a core, what the coverage actually is once duplicates are
gone.

They read a finished database and write tidy CSVs (and, for one module, plots)
under `out_dir/<module>/`. Nothing here **modifies a pipeline database**, so
re-running is always safe.

## Before you start

**`multised_merged.sqlite`** in `db_dir`. That is the whole prerequisite — no
per-source databases, nothing external.

Two modules need a suggested package and name it if missing:

- `clustering` needs **cluster**
- `outlier_review` needs **ggplot2**

## Running them

```{r}
analyze_data("merged", out_dir = "./data/analysis-test")                 # all 13
analyze_data("merged", module = "hotspots", out_dir = "./data/analysis-test")
```

**Pass `out_dir`.** It defaults to the live analysis tree (`data/analysis`), and
a test run left at the default overwrites the files the site is built from.

`db_dir` is read-only here, so pointing it at the real `data/db` is safe, and is
what you want: the analyses should be checked against the database the stored
outputs came from.

About 0.6 minutes for all thirteen — the longest of the three generations, and
still under a minute.

## The thirteen modules

`analysis_modules("merged")` is the executable version of this table.

| Module | Covers | Files |
|---|---|--:|
| `merged_summary` | the **Data Categories** page: every target/reference/organic measurement classified by fraction (bulk / sieved63 / sieved20), covariate availability and layering | 3 |
| `grainsize` | grain size and fines on the merged database | 5 |
| `normalisation` | Fe/Al normalisation | 4 |
| `organic` | organic carbon | 4 |
| `spatial` | depth and distance to coast | 3 |
| `temporal` | sampling year | 4 |
| `depthprofile` | down-core depth profiles | 4 |
| `enrichment` | enrichment | 3 |
| `clustering` | Facies: K-means on chemistry. Needs **cluster** | 4 |
| `hotspots` | Hotspots: DBSCAN on geography | 4 |
| `regions` | Regions: both, weighted | 4 |
| `siteyears` | location-controlled temporal trend, 2dp grid, within-cell | 4 |
| `outlier_review` | review prototype, **not a site input**. Needs **ggplot2** | 5 |

`clustering`, `hotspots` and `regions` are one family rather than three
unrelated modules: chemistry alone, geography alone, and the two combined under
a weight sweep. They make sense read together.

Five module names are shared with the clean generation, and the two write into
the same directory. They do not collide — every file here is prefixed `merged_`
(`merged_grainsize_fines_summary.csv` beside clean's
`grainsize_fines_summary.csv`). See `vignette("analyses-clean")`.

## outlier_review is not a site input

It is the prototype that settled the distributional `outlier_flag` rule the
**merge stage now applies** (`merge_mark_outliers()`, merged step 4). It writes
candidate and summary CSVs plus three distribution plots for eyeballing, and it
runs as part of a full `"merged"` pass, but no Quarto page reads it. The
site's Outlier Flagging page reads the `merge_outlier_*` CSVs written by the
merge step instead.

## What you should get

51 files across thirteen directories. The prefixes make the generation obvious
in a listing:

```
merged_summary/   merged_bulk_factors.csv  merged_coverage_fraction.csv
                  merged_layering.csv
grainsize/        merged_grainsize_*.csv                              (5)
normalisation/    merged_normalisation_*.csv                          (4)
organic/          merged_organic_*.csv                                (4)
spatial/          merged_spatial_*.csv                                (3)
temporal/         merged_temporal_*.csv                               (4)
depthprofile/     merged_depthprofile_*.csv                           (4)
enrichment/       merged_enrichment_*.csv                             (3)
clustering/       merged_clustering_*.csv                             (4)
hotspots/         merged_hotspots_*.csv                               (4)
regions/          merged_regions_*.csv                                (4)
siteyears/        merged_siteyears_*.csv                              (4)
outlier_review/   outlier_candidates.csv  outlier_group_summary.csv
                  outlier_dist_{organic,reference,target}.png
```

`outlier_review` is the one module without the prefix, which is consistent with
it being the one module that feeds no page.

## Checking a rebuild

Deterministic, so given the same database they reproduce byte for byte.

```{r}
compare <- function(out) {
  mods <- basename(list.dirs(out, recursive = FALSE))
  unlist(lapply(mods, function(m) {
    a <- file.path("./data/analysis", m); b <- file.path(out, m)
    f <- intersect(list.files(a), list.files(b))
    stats::setNames(vapply(f, function(x)
      identical(unname(tools::md5sum(file.path(a, x))),
                unname(tools::md5sum(file.path(b, x)))), logical(1)),
      file.path(m, f))
  }))
}

table(compare("./data/analysis-test"))
```

Verified here: **46 of 51 identical**.

### The five that differ

All five are `outlier_review`:

```
outlier_review/outlier_candidates.csv
outlier_review/outlier_group_summary.csv
outlier_review/outlier_dist_organic.png
outlier_review/outlier_dist_reference.png
outlier_review/outlier_dist_target.png
```

**The stored files are stale, not the new ones.** Running the original script
`R/analysis/outlier_review/01_merged_outlier_review.R`, before it was removed at
v0.3.0, into an isolated directory and comparing all three versions by MD5
gives:

| File | converted == original | stored == original |
|---|---|---|
| all five | **TRUE** | FALSE |

So the conversion reproduces the original exactly, and the stored copies match
neither. They predate the merge-stage outlier flagging that settled the rule
this module was written to explore. Re-run the module if you want current
numbers.

Worth knowing before judging a rebuild: five differences that all sit in one
module, and that module a review prototype rather than a site input, is the
expected result — not a conversion defect.

## When it fails

**"unable to open database file"** — `multised_merged.sqlite` is not in
`db_dir`.

**"The clustering analysis needs the suggested package(s): cluster."** — install
it, or run the other twelve modules individually.

**"The outlier_review analysis needs the suggested package(s): ggplot2."** —
same, and this one costs nothing to skip, since no page reads it.

**"The merged generation has no module 'uncertainty'."** — that one is clean
only. `analysis_modules("merged")` lists the thirteen.

**CSVs appearing in `data/analysis`** — `out_dir` was left at its default.

## The site

These outputs feed **multised-merged**. Adding an output means uploading it to
that repo's release before pushing `main`: the CI pre-render downloads the
assets, and a local render hides the problem because the local data symlink is
skip-if-exists.
