---
title: "Analyses: refined"
description: >
  The background suite: one module of six ordered steps on the refined
  database, plus the flat export.
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

The refined generation has **one** analysis module, `background`, and it holds
six steps rather than one. They are the reason the refined mart exists: five
independent ways of asking what a pristine sediment concentration looks like,
and a sixth that puts the answers side by side.

They read a finished database and write tidy CSVs under `out_dir/background/`.
Nothing here **modifies a pipeline database**, so re-running is always safe.

## Before you start

**`multised_refined.sqlite`** in `db_dir`. The aquaculture reference is already
folded into it, so that is the only file needed, and no suggested packages are
involved.

## Running them

```{r}
analyze_data("refined", out_dir = "./data/analysis-test")     # the whole suite
analyze_data("refined", module = "background", steps = 4,
             out_dir = "./data/analysis-test")                # just the EF step
```

**Pass `out_dir`.** It defaults to the live analysis tree (`data/analysis`), and
a test run left at the default overwrites the files multised-refined is built
from.

`db_dir` is read-only here, so pointing it at the real `data/db` is safe, and is
what you want.

About 0.1 minutes — the fastest of the three, because the mart already carries
the ratios the analyses would otherwise compute.

`analyze_data("refined")` reports **one module, not six**. The six are steps
within it, which is why `steps` works here and nowhere else.

## The six steps

| Step | Function | Asks |
|---|---|---|
| 1 | `analysis_refined_background()` | what do the raw percentiles say? |
| 2 | `analysis_refined_background_gsnorm()` | and after grain-size normalisation? |
| 3 | `analysis_refined_background_pressure()` | and if you select on pressure (distance to aquaculture, to coast)? |
| 4 | `analysis_refined_background_ef()` | what does the enrichment factor against Al say? |
| 5 | `analysis_refined_background_mixture()` | can a distribution mixture separate background from contaminated? |
| 6 | `analysis_refined_pristine()` | do the answers agree? |

The six map onto the six background pages of multised-refined, in this order.

### Step 6 depends on 1, 4 and 5

Steps 1-5 are independent: each reads the database and writes its own CSVs, so
any of them can be re-run alone. Step 6 is different — it is a **synthesis**,
and it reads three earlier outputs back out of `out_dir/background/`:

```
refined_background_compare.csv     (step 1)  → the p90 offshore percentile
refined_ef_background.csv          (step 4)  → the EF background ratio
refined_mixture_components.csv     (step 5)  → the mixture threshold
```

So `steps = 6` into a **fresh** `out_dir` fails at the first `read_csv()`. Run
the suite once into that directory first, or point `out_dir` at one that already
has the earlier outputs.

## What you should get

20 files in one directory:

```
background/  refined_background_{compare,meta,percentiles}.csv          (step 1)
             refined_gsnorm_{compare,meta,percentiles}.csv              (step 2)
             refined_pressure_{compare,meta,percentiles}.csv            (step 3)
             refined_ef_{background,dist,meta,pressure}.csv             (step 4)
             refined_mixture_{components,hist,meta}.csv                 (step 5)
             refined_pristine_{coverage,meta,summary,validation}.csv    (step 6)
```

Every step writes a `*_meta.csv` recording the thresholds and filters it used,
which is what makes the six comparable at step 6 rather than merely adjacent.

The headline finding is in `refined_pristine_coverage.csv`, and it is a negative
one worth expecting: **aquaculture-proximate sites cannot be classified for lack
of aluminium** — about 2% Al coverage within 1 km of a farm against 90% beyond
20 km. The enrichment-factor route needs Al, and the sites under most pressure
are the ones that do not have it.

## Checking a rebuild

Deterministic, so given the same database they reproduce byte for byte.

```{r}
a <- "./data/analysis/background"; b <- "./data/analysis-test/background"
f <- intersect(list.files(a), list.files(b))
table(vapply(f, function(x)
  identical(unname(tools::md5sum(file.path(a, x))),
            unname(tools::md5sum(file.path(b, x)))), logical(1)))
```

Verified here: **20 of 20 identical**. There are no known stale outputs in this
generation, so anything `FALSE` is worth investigating.

## The export is not an analysis

The flat downloadable dataset comes from the same database but denormalises
rather than computing, so it lives behind its own verb:

```{r}
export_data("refined", out_dir = "./data/analysis-test")
```

115,820 rows by 16 columns, plus a column dictionary, under
`<out_dir>/download/` — the path the multised-refined pre-render expects.
`refined` is the only generation with an export, and it is deliberately not
reachable as an analysis module: there is exactly one public path to it. See
`export_data()` and `vignette("refined")`.

## When it fails

**"unable to open database file"** — `multised_refined.sqlite` is not in
`db_dir`.

**A `read_csv()` error in step 6** — the earlier outputs are not in
`out_dir/background/`. See above; run steps 1-5 first.

**"The refined generation has no module 'grainsize'."** — the refined generation
has exactly one module. `analysis_modules("refined")` shows its six steps.

**"`steps` selects within one module, so `module` is required with it."** — pass
`module = "background"` alongside `steps`.

**CSVs appearing in `data/analysis`** — `out_dir` was left at its default.

## The site

These outputs feed **multised-refined**, along with the two download pages.
Adding an output means uploading it to that repo's release before pushing
`main`: the CI pre-render downloads the assets, and a local render hides the
problem because the local data symlink is skip-if-exists.
