---
title: "Analyses"
description: >
  The 25 read-only analyses that run on a finished database and write the CSVs
  the four Quarto sites are built from.
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

Analyses read a finished database and write tidy CSVs (and, for two modules,
plots) under `out_dir/<module>/`. They **never modify a pipeline database**, so
re-running one is always safe — the only thing at risk is the output directory.

Each output feeds a page on one of the Quarto sites, which is why adding one
means uploading it to that site repo's release before pushing `main`: the CI
pre-render downloads them, and a local render hides the problem because the
local data symlink is skip-if-exists.

## Running them

```{r}
analyze_data("merged", out_dir = "./data/analysis-test")            # all 13
analyze_data("clean", module = "grainsize", out_dir = "…")          # one module
analyze_data("refined", module = "background", steps = 4, out_dir = "…")
```

**Pass `out_dir`.** It defaults to the live analysis tree, and a test run left at
the default overwrites the files the sites are built from. Nothing is destroyed
that a re-run cannot restore, but there is no reason to.

`db_dir` is read-only here, so pointing it at the real `data/db` is safe and is
what you want: the analyses should be checked against the databases the stored
outputs came from.

Fast. All 25 took under a minute here: clean 0.2, merged 0.6, refined 0.1.

## The 25

`analysis_modules(generation)` is the executable version of these tables.

**clean** — 6 modules, per source, feeding multised-clean:

`grainsize`, `normalisation`, `organic`, `spatial`, `temporal`, `uncertainty`

**merged** — 13 modules, feeding multised-merged:

`merged_summary`, `grainsize`, `normalisation`, `organic`, `spatial`,
`temporal`, `depthprofile`, `enrichment`, `clustering`, `hotspots`, `regions`,
`siteyears`, `outlier_review`

**refined** — `background`, one module holding **6 ordered steps**, mapping onto
the six background pages of multised-refined. `analyze_data("refined")` therefore
reports one module, not six.

Note that `grainsize`, `normalisation`, `organic`, `spatial` and `temporal`
exist for both clean and merged, and **share an output directory**. That is why
the token in the function name is the generation (`analysis_clean_grainsize()`
vs `analysis_merged_grainsize()`) rather than the module alone, and why a
directory listing mixes outputs from two generations.

Two modules need a suggested package and say so by name if it is missing:
`clustering` needs **cluster**, `outlier_review` needs **ggplot2**.

## What you should get

File counts per module directory, after all three generations have run:

| Module | Files | | Module | Files |
|---|--:|---|---|--:|
| `background` | 20 | | `merged_summary` | 3 |
| `clustering` | 4 | | `normalisation` | 8 |
| `depthprofile` | 4 | | `organic` | 8 |
| `enrichment` | 3 | | `outlier_review` | 5 |
| `grainsize` | 10 | | `regions` | 4 |
| `hotspots` | 4 | | `siteyears` | 4 |
| `spatial` | 6 | | `temporal` | 7 |
| `uncertainty` | 5 | | | |

95 files in total that also exist in the stored tree.

## Checking a rebuild

Analyses are deterministic, so given the same database they reproduce byte for
byte.

```{r}
compare <- function(out) {
  mods <- basename(list.dirs(out, recursive = FALSE))
  unlist(lapply(mods, function(m) {
    a <- file.path("./data/analysis", m); b <- file.path(out, m)
    f <- intersect(list.files(a), list.files(b))
    stats::setNames(vapply(f, function(x)
      identical(unname(tools::md5sum(file.path(a, x))),
                unname(tools::md5sum(file.path(b, x)))), logical(1)),
      file.path(m, f))
  }))
}

table(compare("./data/analysis-test"))
```

Verified here: **90 of 95 identical**.

### The five that differ

All five are `outlier_review`:

```
outlier_review/outlier_candidates.csv
outlier_review/outlier_group_summary.csv
outlier_review/outlier_dist_organic.png
outlier_review/outlier_dist_reference.png
outlier_review/outlier_dist_target.png
```

**The stored files are stale, not the new ones.** Running the original script
`R/analysis/outlier_review/01_merged_outlier_review.R` into an isolated directory
and comparing all three versions by MD5 gives:

| File | converted == original | stored == original |
|---|---|---|
| all five | **TRUE** | FALSE |

So the conversion reproduces the original exactly, and the stored copies match
neither. They predate the merge-stage outlier flagging (`merge_mark_outliers()`,
merged step 4) that settled the rule this module was written to explore. Re-run
the module if you want current numbers.

This is worth knowing before judging a rebuild: five differences that all sit in
one module, with the module that produced them being a review prototype rather
than a site input, is the expected result — not a conversion defect.

## outlier_review is not a site input

It is the prototype that settled the distributional `outlier_flag` rule the merge
stage now applies. It writes candidate and summary CSVs plus distribution plots
for eyeballing, and it is included in a full `"merged"` run, but no Quarto page
reads it.

## The export is not an analysis

The flat downloadable dataset denormalises rather than computing, so it lives
behind its own verb:

```{r}
export_data("refined", out_dir = "./data/analysis-test")
```

It still writes to `<out_dir>/download/`, the path multised-refined's pre-render
expects. It is deliberately not reachable as an analysis module — there is
exactly one public path to it.

## When it fails

**"The refined generation has no module 'grainsize'."** — module names are
validated against the generation. `analysis_modules("refined")` lists what is
available.

**"`steps` selects within one module, so `module` is required with it."** —
only `background` has more than one step.

**"The clustering analysis needs the suggested package(s): cluster."** — install
it, or run the other twelve merged modules individually.

**CSVs appearing in `data/analysis`** — `out_dir` was left at its default.
