---
title: "Analyses: clean"
description: >
  Six per-source analyses on the five clean databases, feeding the
  multised-clean site.
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

The clean analyses answer *what is in each source, and how comparable is it?*
They run before any cross-source merging, which is the point: results are tagged
by source and kept apart, so a problem that belongs to Vannmiljø stays visible
as Vannmiljø's rather than being averaged into a total.

They read a finished database and write tidy CSVs under `out_dir/<module>/`.
Nothing here **modifies a pipeline database**, so re-running is always safe; the
only thing at risk is the output directory.

## Before you start

**All five clean databases** in `db_dir`. Each module loops over the five stems
unconditionally, so a missing one is an error rather than a smaller result.

Nothing external, and no suggested packages: the clean analyses read SQLite and
write CSVs.

## Running them

```{r}
analyze_data("clean", out_dir = "./data/analysis-test")                  # all 6
analyze_data("clean", module = "grainsize", out_dir = "./data/analysis-test")
```

**Pass `out_dir`.** It defaults to the live analysis tree (`data/analysis`),
which is what the multised-clean pre-render reads, so a test run left at the
default overwrites the site's inputs. Nothing is destroyed that a re-run cannot
restore, but there is no reason to.

`db_dir` is read-only here, so pointing it at the real `data/db` is safe, and is
what you want: the analyses should be checked against the databases the stored
outputs came from.

About 0.2 minutes for all six.

## The six modules

`analysis_modules("clean")` is the executable version of this table.

| Module | Covers | Files |
|---|---|--:|
| `grainsize` | sediment fraction and <63µm fines behind the target concentrations | 5 |
| `normalisation` | Fe/Al normalisation | 4 |
| `organic` | organic carbon | 4 |
| `spatial` | depth and distance to coast | 3 |
| `temporal` | sampling year | 3 |
| `uncertainty` | measurement uncertainty | 5 |

Each module writes one set of CSVs covering **all five sources**, with a
`Source` column, rather than one file per source. Depths are kept in full but
reported by the EFSA depth bands (0-5 / 5-40 / >40 cm), assigned from the
subsample interval midpoint.

### Sharing directories with merged

`grainsize`, `normalisation`, `organic`, `spatial` and `temporal` **also exist
for the merged generation**, and the two write into the same module directory.
They do not collide: every merged file is prefixed `merged_`
(`merged_grainsize_fines_summary.csv` beside `grainsize_fines_summary.csv`).

This is why the generation token is in the function name
(`analysis_clean_grainsize()` vs `analysis_merged_grainsize()`) rather than the
module name alone, and why a directory listing mixes the outputs of two
generations. See `vignette("analyses-merged")`.

## What you should get

24 files across six directories:

```
grainsize/      grainsize_bulk_vs_sieved.csv  grainsize_conc_vs_fines.csv
                grainsize_fines_summary.csv   grainsize_fraction_summary.csv
                grainsize_targets_fines.csv
normalisation/  normalisation_availability.csv  normalisation_correlation.csv
                normalisation_pairs.csv         normalisation_ratios.csv
organic/        organic_availability.csv  organic_correlation.csv
                organic_distribution.csv  organic_pairs.csv
spatial/        spatial_enrichment.csv  spatial_normaliser.csv  spatial_pairs.csv
temporal/       temporal_enrichment.csv  temporal_normaliser.csv  temporal_pairs.csv
uncertainty/    uncertainty_by_element.csv  uncertainty_by_method.csv
                uncertainty_coverage.csv    uncertainty_pairs.csv
                uncertainty_rel_values.csv
```

All 24 also exist in the stored tree.

## Checking a rebuild

Deterministic, so given the same databases they reproduce byte for byte.

```{r}
compare <- function(out, prefix = "^(?!merged_)") {
  mods <- basename(list.dirs(out, recursive = FALSE))
  unlist(lapply(mods, function(m) {
    a <- file.path("./data/analysis", m); b <- file.path(out, m)
    f <- intersect(list.files(a), list.files(b))
    f <- grep(prefix, f, perl = TRUE, value = TRUE)
    stats::setNames(vapply(f, function(x)
      identical(unname(tools::md5sum(file.path(a, x))),
                unname(tools::md5sum(file.path(b, x)))), logical(1)),
      file.path(m, f))
  }))
}

table(compare("./data/analysis-test"))
```

Verified here: **24 of 24 identical**. Unlike the merged generation, the clean
analyses have no known stale outputs, so anything `FALSE` is worth
investigating.

## When it fails

**"unable to open database file"** — one of the five clean databases is not in
`db_dir`. All five, or none; there is no per-source clean analysis.

**"The clean generation has no module 'clustering'."** — module names are
validated against the generation. `analysis_modules("clean")` lists the six.

**"`steps` selects within one module, so `module` is required with it."** — no
clean module has more than one step, so `steps` has nothing to select here.

**CSVs appearing in `data/analysis`** — `out_dir` was left at its default.

## The site

These outputs feed **multised-clean**, which also carries the aquaculture and
merge-build pages. Adding an output means uploading it to that repo's release
before pushing `main`: the CI pre-render downloads the assets, and a local
render hides the problem because the local data symlink is skip-if-exists.
