---
title: "Generation 4: merged"
description: >
  Union the five clean databases into one, remove cross-source duplicates,
  finalise the keys and flag distributional outliers.
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

The first cross-source database. Five clean databases go in, one
`multised_merged.sqlite` comes out, with the same measurement rows minus those
that were the same measurement reported twice by two providers.

It takes no `source`. Passing one is an error, not a filter.

## Before you start

**All five clean databases** in `db_dir`. Merged does not build them, and it
will not merge a subset: a missing source is a missing source, not a smaller
union.

**Nothing external.** No seastamp, no `sf` — the location attributes are
carried over from clean, not recomputed. Whatever the clean databases hold is
what the merged database inherits, which matters for the comparison below.

## Running it

```{r}
create_db("merged",
          db_dir  = "./data/multised-test",
          out_dir = "./data/multised-test/analysis")
```

**Pass `out_dir`.** Three of the five steps write analysis CSVs as well as
touching the database, and `out_dir` defaults to the live analysis tree
(`data/analysis`), where the site pre-render steps read from. A test rebuild
that leaves it at the default overwrites those files. They regenerate
identically from the same inputs, so it is recoverable rather than fatal, but
there is no reason to find out.

About half a minute for the whole thing.

### Steps

| Step | Name | Does | Writes CSVs |
|---|---|---|:-:|
| 1 | union | stack all five, prefixing keys per source | |
| 2 | dedup | remove cross-source duplicates | ✓ |
| 3 | finalise | renumber to clean surrogate keys | |
| 4 | mark_outliers | `outlier_flag` on `measurement` | ✓ |
| 5 | summary | retention and stage-total reporting | ✓ |

Steps run in sequence on a database step 1 rebuilds, so a subset only makes
sense once the earlier steps have run. Step 5 computes nothing and can be
re-run freely.

## Deduplication

The same sediment sample can reach two providers — a Norwegian cruise reported
both nationally and to ICES, say. Duplicates are matched on position, date,
depth and element, and resolved by a fixed source preference:

**Mareano > 4Demon > MUDAB > Vannmiljø > ICES-DOME**

National archives beat the international aggregator, since the aggregator is
usually the copy. The rule and its 1% exact-date threshold are in
[merge-pipeline.md](https://github.com/seafood-hazards/multised-engine/blob/main/docs/merge-pipeline.md).

The retention this produces is the most informative single number in the whole
generation:

```
    Source clean_n merged_n retained_pct
   Mareano   24098    24098        100.0
    4Demon    4459     4459        100.0
     MUDAB   30744    30744        100.0
 Vannmiljø   82011    81355         99.2
 ICES-DOME   69719    50188         72.0
```

ICES-DOME loses 28% because it is the aggregator: those rows are held by a
national source too. The three sources above it in the preference order lose
nothing by construction.

## What you should get

The stored `multised_merged.sqlite`:

| Table | Rows |
|---|--:|
| `dataset` | 78 |
| `element` | 69 |
| `site` | 26,849 |
| `event` | 46,091 |
| `subsample` | 53,143 |
| `measurement` | 190,844 |
| `method` | 1,485 |
| `grain_size_fraction` | 104,804 |

Stage totals, which say what each step removed:

| Table | clean total | final | removed | retained % |
|---|--:|--:|--:|--:|
| `dataset` | 78 | 78 | 0 | 100.0 |
| `event` | 46,680 | 46,091 | 589 | 98.7 |
| `grain_size_fraction` | 104,804 | 104,804 | 0 | 100.0 |
| `measurement` | 211,031 | 190,844 | 20,187 | 90.4 |
| `method` | 1,642 | 1,485 | 157 | 90.4 |
| `site` | 27,015 | 26,849 | 166 | 99.4 |
| `subsample` | 53,747 | 53,143 | 604 | 98.9 |

Outliers, from step 4: 381 high, 272 low, 190,191 in distribution.

## Checking a rebuild: mind the inputs

Merged is deterministic, so it reproduces exactly **given the same clean
databases**. That last clause is the whole difficulty.

If you merge the clean databases you rebuilt yourself, they carry
`region = "auto"` geo columns while the stored merged database carries
`"global"` ones, so `dist_to_coast`, `municipality` and `country` will differ —
inherited from clean, not introduced here. Row counts and every chemistry column
should still match exactly, since dedup keys on position, date, depth and
element, and 3-decimal-place positions are not moved by the projection change.

To test the merge logic on its own, merge the **stored** clean databases:

```{r}
dir.create("./data/merge-test")
file.copy(Sys.glob("./data/db/*_clean.sqlite"), "./data/merge-test")

create_db("merged", db_dir = "./data/merge-test",
          out_dir = "./data/merge-test/analysis")
```

That was verified here: all six analysis CSVs byte-identical to the live ones,
and the row counts above reproduced.

```{r}
a <- DBI::dbConnect(RSQLite::SQLite(), "./data/db/multised_merged.sqlite")
b <- DBI::dbConnect(RSQLite::SQLite(), "./data/merge-test/multised_merged.sqlite")
vapply(DBI::dbListTables(a), function(t)
  identical(DBI::dbReadTable(a, t), DBI::dbReadTable(b, t)), logical(1))
```

## When it fails

**"The merged generation combines every source, so `source` must be NULL."** —
there is no per-source merge.

**"unable to open database file"** — one of the five clean databases is not in
`db_dir`. All five, or none.

**CSVs appearing in `data/analysis/merge`** — `out_dir` was left at its default.
Nothing is lost that a re-run does not restore, but pass it.

**Retention that is not 100% for Mareano, 4Demon or MUDAB** — those three sit
above Vannmiljø and ICES-DOME in the preference order and cannot lose rows to
them. If they do, the preference order or the matching key has changed.

## Next

`create_db("refined")` cuts a mart from this database for the pristine and
background work: one measurement table, a slim normaliser, baked ratios, and the
aquaculture and repeat-site links.
