---
title: "Generation 1: pilot"
description: >
  Parse each vendor's raw export into a per-source database. What each source
  needs, what to run, and what the result should look like.
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

The pilot generation reads each data provider's own files and lands them in
SQLite with as little interpretation as possible. It is the only stage that
touches the vendor formats, and the only one whose table structure differs per
source: Mareano has 6 tables, Vannmiljø 10, and the names have nothing in
common. Harmonisation is the *next* generation's job.

Nothing in this article can be run from a checkout alone: every step needs the
raw exports, which are not distributed with the package.

## Before you start

**The raw exports**, under `data/raw/<Source>/`. Each parser checks its inputs
up front and names any that are missing, so a wrong path fails immediately
rather than half way through a parse.

| Source | Directory | Files |
|---|---|---|
| `mareano` | `Mareano/` | `Mareano.xlsx`, `P2301_surfacedata_GISprepared_draft.xlsx`, `P2501_surfacesamples_GISprepared_ICP_coulter_POPs.xlsx` |
| `vannmiljo` | `Vannmiljo/` | `Vannmilio_Elements_interest.xlsx`, `Vannmilio_Elements_others.xlsx`, `Vannmilio_pH_Carbon_Sulfur_all.xlsx`, `Vannmilio_Partikler_all.xlsx` |
| `ices-dome` | `ICES DOME/` | `DomeSediment_Data_0326015962.csv` plus the `code/` lookup directory |
| `mudab` | `Mudab/` | `export_mudab_V_MESSWERTE_SEDIMENT.csv`, `export_mudab.tsv`, `MUDAB_CODELISTEN.tsv` |
| `4demon` | `4Demon/` | `20200713_1007454DEMON_HeavyMetals_PCBs_in_sediment.csv` |

Note Vannmiljø's filenames say `Vannmilio` while the directory says `Vannmiljo`.
That is how the exports arrived; it is not a typo to fix.

**Suggested R packages**, needed by two sources only:

- `readxl` — Mareano and Vannmiljø (Excel inputs)
- `sf` — Vannmiljø only, to reproject UTM33 to longitude/latitude

`sf` needs GDAL, GEOS, PROJ and udunits at the system level. If it installs but
will not load, the libraries arrived after `sf` did; rebuild with
`renv::install("sf", rebuild = TRUE)`.

**The seastamp CLI** and its reference datasets, for step 4. See
`vignette("clean")` for the dataset table; the short version is five downloads
totalling about 7.7 GB, of which GEBCO is 7 GB. Skipping step 4 is a supported
option (below).

## Running it

```{r}
library(multised.engine)

create_db("pilot", "mareano")
```

That is the whole command. It writes `data/db/mareano_pilot.sqlite` and prints a
summary per step.

Two arguments are worth knowing:

```{r}
# build somewhere harmless while testing, rather than over the real databases
create_db("pilot", "mareano", db_dir = "./data/multised-test")

# point at seastamp explicitly: the RStudio console does not inherit the login
# shell's PATH, so a binary the Terminal tab can see is often invisible here
create_db("pilot", "mareano", seastamp_bin = "/path/to/seastamp")
```

`db_dir` defaults to `multised_db_dir()` (`data/db`), so leaving it out writes
to the real working set. While validating, set it once for the session:

```{r}
options(multised.db_dir = "./data/multised-test")
```

The five sources are independent, so order does not matter:

```{r}
for (src in multised_sources()) create_db("pilot", src)
```

### Steps

The step numbers are inherited from the original scripts, which is why they are
1, 4, 5 rather than 1, 2, 3.

| Step | Does |
|---|---|
| 1 | parse the vendor files into data frames |
| 4 | annotate the station frame with seastamp |
| 5 | create the schema and write the tables |

Steps 4 and 5 consume what step 1 builds, so step 1 cannot be skipped; asking
for `steps = 5` is an error rather than a silent no-op.

**Without seastamp**, run the parse and the write only:

```{r}
create_db("pilot", "mareano", steps = c(1, 5))
```

The six location columns come out `NULL`. That costs nothing downstream: clean
step 4 recomputes all of them anyway, and those are the values that survive into
the merged database. Pilot geo-enrichment exists so the pilot databases are
inspectable on their own, not because anything later depends on it.

## What you should get

Row counts of the five stored pilot databases. A rebuild should reproduce these
exactly; if a number moves, the raw export changed or something is wrong.

| Source | Tables | Rows |
|---|--:|---|
| `mareano` | 6 | core 404, cruise 40, lld 994, parameter 60, sample 3553, sediment 144040 |
| `vannmiljo` | 10 | activity 31, analysis_method 31, client 1289, contractor 307, lld 54952, parameter 50, sample 45852, sample_method 9, sediment 238254, site 24330 |
| `ices-dome` | 9 | analysis_method 2609, code_lookup 708, lld 1536, parameter 142, project 148, reference 27, sample 17694, sediment 345809, site 11306 |
| `mudab` | 10 | analysis_method 342, code_lookup 279, lod 482, measurement_time 5514, parameter 61, reference_material 542, sample 14089, sediment 167124, station 823, survey 3981 |
| `4demon` | 6 | method 176, parameter 17, project 222, sample 1796, sediment 23542, station 166 |

The station frame differs per source, which is also where the geo columns land:
`core` for Mareano, `site` for Vannmiljø and ICES-DOME, `survey` for MUDAB,
`station` for 4Demon. Mareano is the one source that names the nearest-country
column `country`; the others call it `est_country`.

Console output for a full Mareano run ends like this:

```
== mareano pilot step 1: extract ==
mareano parsed: cruise 40, core 404, sample 3553, parameter 60, sediment 144040, lld 994

== mareano pilot step 4: geo (seastamp) ==
seastamp: 394 points annotated (region auto)
  394 distinct positions -> 404 station rows annotated

== mareano pilot step 5: write ==
  core 404 | cruise 40 | lld 994 | parameter 60 | sample 3553 | sediment 144040
```

seastamp runs over *distinct* coordinates, so the two numbers differ wherever
stations repeat a position: 394 positions for 404 Mareano cores, and 24,193 for
24,330 Vannmiljø sites. That also sets the runtime — Mareano's geo step is
seconds, Vannmiljø's is minutes.

## Checking a rebuild

Compare the table row counts against the table above, and the measurement
columns against the previous database. The counts are the strong signal: they
are large, source-specific, and a parsing change moves them.

```{r}
con <- DBI::dbConnect(RSQLite::SQLite(), "./data/multised-test/mareano_pilot.sqlite")
sapply(DBI::dbListTables(con), function(t)
  DBI::dbGetQuery(con, paste0("SELECT COUNT(*) FROM ", t))[[1]])
DBI::dbDisconnect(con)
```

**Do not compare the geo columns against a stored pilot database.** They will
differ, by design: those values came from an earlier implementation using `sf`
with rnaturalearth and giscoR, which seastamp replaced. Same reference datasets,
different code, so the numbers disagree in detail. Judge them on whether they
are plausible — distances positive and sane, countries correct, `sea_name`
populated — not on whether they reproduce.

The `region` argument compounds this: the code uses seastamp's own
`region = "auto"`, while the stored databases hold `"global"` values.
`dist_to_coast` moves on every row between the two (median 4.8-11.3% by source,
largest single shift 93 km), `municipality` for 1,884 sites and `country` for
283. `depth` does not project and is unchanged.

## When it fails

**"Mareano export not found: …"** — the input check. `create_db()` has no
`raw_dir` argument, so redirect with the option:

```{r}
options(multised.raw_dir = "/path/to/raw")
```

**"The Vannmiljo pilot parser needs the suggested package(s): sf."** — either
`sf` is absent, or it is installed and will not load. Run `library(sf)` to see
which; the message says so too. `renv::restore()` fixes the first, system
libraries plus a rebuild fix the second.

**"seastamp not found on the PATH."** — resolution order is
`options(multised.seastamp_bin)`, then `SEASTAMP_BIN`, then the `PATH`. In
RStudio prefer one of the first two, since the console does not inherit the
shell's `PATH`. Or skip step 4 with `steps = c(1, 5)`.

**"seastamp reference data not found for: …"** — the datasets, not the binary.
All five are validated up front because the four seastamp commands run as one
chain, so even a run that only wants `dist_to_coast` needs GEBCO present.

**A long wall of `` • `` -> `...227` `` messages** while an Excel sheet loads.
That is readxl repairing empty column names, and it is expected for Mareano and
Vannmiljø.

## Next

`create_db("slim", source)` reads `<source>_pilot.sqlite` and reshapes it into
the shared seven-table schema. From there the generations are common across
sources, and the per-source differences are down to a handful of columns.
