---
title: "Generation 1: pilot"
description: >
  Parse each vendor's raw export into a per-source database. What each source
  needs, what to run, and what the result should look like.
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

The pilot generation reads each data provider's own files and lands them in
SQLite with as little interpretation as possible. It is the only stage that
touches the vendor formats, and the only one whose table structure differs per
source: Mareano has 6 tables, Vannmiljø 10, and the names have nothing in
common. Harmonisation is the *next* generation's job.

Nothing in this article can be run from a checkout alone: every step needs the
raw exports, which are not distributed with the package.

## Before you start

**The raw exports**, under `data/raw/<Source>/`. Each parser checks its inputs
up front and names any that are missing, so a wrong path fails immediately
rather than half way through a parse.

| Source | Directory | Files |
|---|---|---|
| `mareano` | `Mareano/` | `Mareano.xlsx`, `P2301_surfacedata_GISprepared_draft.xlsx`, `P2501_surfacesamples_GISprepared_ICP_coulter_POPs.xlsx` |
| `vannmiljo` | `Vannmiljo/` | `Vannmilio_Elements_interest.xlsx`, `Vannmilio_Elements_others.xlsx`, `Vannmilio_pH_Carbon_Sulfur_all.xlsx`, `Vannmilio_Partikler_all.xlsx` |
| `ices-dome` | `ICES DOME/` | `DomeSediment_Data_0326015962.csv` plus the `code/` lookup directory |
| `mudab` | `Mudab/` | `export_mudab_V_MESSWERTE_SEDIMENT.csv`, `export_mudab.tsv`, `MUDAB_CODELISTEN.tsv` |
| `4demon` | `4Demon/` | `20200713_1007454DEMON_HeavyMetals_PCBs_in_sediment.csv` |

Note Vannmiljø's filenames say `Vannmilio` while the directory says `Vannmiljo`.
That is how the exports arrived; it is not a typo to fix.

**Suggested R packages**, needed by two sources only:

- `readxl` — Mareano and Vannmiljø (Excel inputs)
- `sf` — Vannmiljø only, to reproject UTM33 to longitude/latitude

`sf` needs GDAL, GEOS, PROJ and udunits at the system level. If it installs but
will not load, the libraries arrived after `sf` did; rebuild with
`renv::install("sf", rebuild = TRUE)`.

**No seastamp.** The pilot stage needs neither the CLI nor its 7.7 GB of
reference data. It declares the five location columns and leaves them empty;
clean step 4 fills them. See `vignette("clean")`.

## Running it

```{r}
library(multised.engine)

create_db("pilot", "mareano")
```

That is the whole command. It writes `data/db/mareano_pilot.sqlite` and prints a
summary per step.

Two arguments are worth knowing:

```{r}
# build somewhere harmless while testing, rather than over the real databases
create_db("pilot", "mareano", db_dir = "./data/multised-test")
```

`db_dir` defaults to `multised_db_dir()` (`data/db`), so leaving it out writes
to the real working set. While validating, set it once for the session:

```{r}
options(multised.db_dir = "./data/multised-test")
```

The five sources are independent, so order does not matter:

```{r}
for (src in multised_sources()) create_db("pilot", src)
```

### Steps

The step numbers are inherited from the original scripts, which is why they are
1 and 5 rather than 1 and 2.

| Step | Does |
|---|---|
| 1 | parse the vendor files into data frames |
| 5 | create the schema and write the tables |

Step 5 consumes what step 1 builds, so step 1 cannot be skipped; asking for
`steps = 5` is an error rather than a silent no-op.

**There was a step 4**, which ran seastamp over every station exactly as clean
step 4 does. It was removed on 2026-08-25: `clean_geo_enrich()` UPDATEs all five
location columns unconditionally, so no pilot value ever survived. Computing
them twice bought nothing and cost a second seastamp dependency and a second set
of numbers that could disagree with the ones that do survive. The columns stay,
because the pilot schema declares them and the slim transforms select them; only
the values go. See `R/pilot-geo.R`, and git history for the implementation if
the pilot values are ever wanted back.

## What you should get

Row counts of the five stored pilot databases. A rebuild should reproduce these
exactly; if a number moves, the raw export changed or something is wrong.

| Source | Tables | Rows |
|---|--:|---|
| `mareano` | 6 | core 404, cruise 40, lld 994, parameter 60, sample 3553, sediment 144040 |
| `vannmiljo` | 10 | activity 31, analysis_method 31, client 1289, contractor 307, lld 54952, parameter 50, sample 45852, sample_method 9, sediment 238254, site 24330 |
| `ices-dome` | 9 | analysis_method 2609, code_lookup 708, lld 1536, parameter 142, project 148, reference 27, sample 17694, sediment 345809, site 11306 |
| `mudab` | 10 | analysis_method 342, code_lookup 279, lod 482, measurement_time 5514, parameter 61, reference_material 542, sample 14089, sediment 167124, station 823, survey 3981 |
| `4demon` | 6 | method 176, parameter 17, project 222, sample 1796, sediment 23542, station 166 |

The station frame differs per source, which is also where the empty location
columns sit: `core` for Mareano, `site` for Vannmiljø and ICES-DOME, `survey`
for MUDAB, `station` for 4Demon. Mareano is the one source that names the
nearest-country column `country`; the others call it `est_country`.

Console output for a full Mareano run ends like this:

```
== mareano pilot step 1: extract ==
mareano parsed: cruise 40, core 404, sample 3553, parameter 60, sediment 144040, lld 994

== mareano pilot step 5: write ==
  core 404 | cruise 40 | lld 994 | parameter 60 | sample 3553 | sediment 144040
```

## Checking a rebuild

Compare the table row counts against the table above, and the measurement
columns against the previous database. The counts are the strong signal: they
are large, source-specific, and a parsing change moves them.

```{r}
con <- DBI::dbConnect(RSQLite::SQLite(), "./data/multised-test/mareano_pilot.sqlite")
sapply(DBI::dbListTables(con), function(t)
  DBI::dbGetQuery(con, paste0("SELECT COUNT(*) FROM ", t))[[1]])
DBI::dbDisconnect(con)
```

**The whole database reproduces**, because nothing in it is derived from an
external service any more. The five location columns are present and `NULL` in
all five stored pilot databases, on all 404 Mareano cores, 24,330 Vannmiljø
sites, 11,306 ICES-DOME sites, 3,981 MUDAB surveys and 166 4Demon stations.
Where those values come from, which seastamp version produced them and how they
move under a different projection is entirely a clean-stage question: see
`vignette("clean")` and [clean-pipeline.md](https://github.com/seafood-hazards/multised-engine/blob/main/docs/clean-pipeline.md).

## When it fails

**"Mareano export not found: …"** — the input check. `create_db()` has no
`raw_dir` argument, so redirect with the option:

```{r}
options(multised.raw_dir = "/path/to/raw")
```

**"The Vannmiljo pilot parser needs the suggested package(s): sf."** — either
`sf` is absent, or it is installed and will not load. Run `library(sf)` to see
which; the message says so too. `renv::restore()` fixes the first, system
libraries plus a rebuild fix the second.

**Anything about seastamp** — it is not the pilot stage. `create_db()` accepts
`seastamp_dir` and `seastamp_bin` for every generation, but only clean step 4
reads them, so a pilot run cannot fail on either. See `vignette("clean")`.

**A long wall of `` • `` -> `...227` `` messages** while an Excel sheet loads.
That is readxl repairing empty column names, and it is expected for Mareano and
Vannmiljø.

## Publishing to the pilot sites

Each source has its own Quarto site — `mareano-pilot`, `vannmiljo-pilot`,
`ices-dome-pilot`, `mudab-pilot`, `4demon-pilot` — and each depends on exactly
one file: the `<source>_pilot.sqlite` this generation writes. The site downloads
it from its own repository's **latest** release, so a rebuilt database reaches
the site by being attached to a new release there:

```{r}
# from the site repo, after tagging vX.Y.Z
# gh release create vX.Y.Z --title vX.Y.Z --notes-from-tag \
#   ../multised-engine/data/db/mareano_pilot.sqlite
```

Two things to know before doing that:

- **`releases/latest/download/` does not fall back.** A release published without
  the database as an asset breaks that site's next render, so every release must
  carry it, and the release must exist before `main` is pushed.
- **The content change needs a `cacheKey` bump** in the site's `_db-setup.qmd`,
  or returning browsers keep serving the cached old copy.

The site repos' `docs/site.md` files carry the full procedure. See
[websites.md](https://github.com/seafood-hazards/multised-engine/blob/main/docs/websites.md)
for how the five pilot sites differ from the four generation sites.

## Next

`create_db("slim", source)` reads `<source>_pilot.sqlite` and reshapes it into
the shared seven-table schema. From there the generations are common across
sources, and the per-source differences are down to a handful of columns.
