---
title: "Generation 5: refined"
description: >
  A mart cut from the merged database for the pristine and background work:
  target elements only, normalisers reshaped, ratios baked in.
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

Refined is not another harmonisation pass. The merged database is already the
complete, corrected record, and stays that way; refined is a **mart** cut from
it for one job — deciding what a pristine sediment concentration looks like.

That job wants a particular shape. Every target measurement in one flat-ish
table, its normalisers beside it rather than as sibling rows, the ratios already
divided, and the pressure indicators (aquaculture proximity, repeat sampling)
attached to the site. Computing those on the fly across 190,000 rows in six
analyses was the alternative.

Takes no `source`.

## Before you start

**`multised_merged.sqlite`** in `db_dir`, and **`aquaculture_no.sqlite`** for
step 4. `sf` is needed by step 4 only.

## What changes from merged

The reshaping is the point, so it is worth stating plainly. The merged
`measurement` table holds every measurand as a row. Refined splits them by role:

```
merged measurement            190,844
  ├─ 7 target elements        115,820  → measurement
  ├─ FE / AL / CORG            62,698  → normaliser (reshaped, one row per subsample)
  └─ TOC63                     12,326  → dropped
```

TOC63 goes because it is fines-normalised organic carbon, not whole-sample
CORG, so it is not comparable with the CORG the ratios use. The three numbers
add back to 190,844 exactly, which is what `refined_reconciliation.csv` records
— nothing vanishes unaccounted for.

`measurement` then carries `ratio_fe`, `ratio_al` and `ratio_corg` already
divided, and `element` shrinks to the 7 targets.

| Table | Rows |
|---|--:|
| `measurement` | 115,820 |
| `normaliser` | 38,145 |
| `subsample` | 53,143 |
| `event` | 46,091 |
| `site` | 26,849 |
| `method` | 949 |
| `dataset` | 78 |
| `element` | **7** |
| `aquaculture` | 3,743 |

The aquaculture reference is **copied into** the refined database rather than
referenced, so the mart is self-contained: one file holds everything the
background analyses need.

## Running it

```{r}
create_db("refined",
          db_dir  = "./data/multised-test",
          out_dir = "./data/multised-test/analysis")
```

Fast — about 6 seconds. Step 6 writes CSVs, so pass `out_dir` or they land in
the live analysis tree.

| Step | Name | Does |
|---|---|---|
| 1 | restructure | cut the mart: targets to `measurement`, normalisers reshaped |
| 2 | normaliser | fill the `normaliser` table per subsample |
| 3 | ratios | bake `ratio_fe`, `ratio_al`, `ratio_corg` |
| 4 | aquaculture | nearest-farm link on `site`; needs `sf` |
| 5 | repeat_sites | `repeat_group` and `n_years` per location cell |
| 6 | summary | reconciliation and coverage CSVs |

## What you should get

Step 6 prints the coverage table, which is the honest picture of what the
background work has to live with:

| Scope | Field | Present | Total | % |
|---|---|--:|--:|--:|
| normaliser bulk | fe | 8,487 | 27,704 | 31 |
| normaliser bulk | al | 9,193 | 27,704 | 33 |
| normaliser bulk | corg | 24,542 | 27,704 | 89 |
| normaliser sieved | fe | 6,556 | 10,441 | 63 |
| normaliser sieved | al | 8,992 | 10,441 | 86 |
| normaliser sieved | corg | 4,112 | 10,441 | 39 |
| measurement bulk | ratio_al | 30,647 | 86,018 | 36 |
| measurement sieved | ratio_al | 23,256 | 29,802 | 78 |
| site | aqua_id | 18,696 | 26,849 | 70 |
| site | repeat n_years>=3 | 10,317 | 26,849 | 38 |

Aluminium on bulk samples is the binding constraint at 33%, and it is the
normaliser the enrichment-factor work depends on. That is a property of the
sources, not of this generation, but this is where it becomes visible.

## Checking a rebuild

Refined is deterministic and touches nothing external except the aquaculture
reference, so given the same merged database it reproduces **exactly**.

```{r}
dir.create("./data/refine-test")
file.copy(c("./data/db/multised_merged.sqlite", "./data/db/aquaculture_no.sqlite"),
          "./data/refine-test")

create_db("refined", db_dir = "./data/refine-test",
          out_dir = "./data/refine-test/analysis")

a <- DBI::dbConnect(RSQLite::SQLite(), "./data/db/multised_refined.sqlite")
b <- DBI::dbConnect(RSQLite::SQLite(), "./data/refine-test/multised_refined.sqlite")
vapply(DBI::dbListTables(a), function(t)
  identical(DBI::dbReadTable(a, t), DBI::dbReadTable(b, t)), logical(1))
```

Verified here: **9 of 9 tables identical**, and all three step-6 CSVs identical
too. Copying the stored merged database first is what makes that exact — merging
your own clean databases first would carry their `region = "auto"` geo columns
through into `site`.

## The export

```{r}
export_data("refined", out_dir = "./data/multised-test/analysis")
```

The flat downloadable dataset comes from this database: 115,820 rows by 16
columns, plus a column dictionary, both under `<out_dir>/download/`. It is a
separate verb rather than an analysis module because it denormalises and
computes nothing. See `export_data()`.

## When it fails

**"The refined generation combines every source, so `source` must be NULL."**

**"unable to open database file"** — `multised_merged.sqlite` is not in `db_dir`.

**"The refined aquaculture step needs the suggested package(s): sf."** — step 4
only. `steps = c(1:3, 5:6)` gets everything else, leaving `aqua_id` NULL.

**Reconciliation that does not add up** — `refined_reconciliation.csv` should
account for every merged measurement row as target, normaliser or dropped
TOC63. If the three do not sum to the merged total, something was lost silently.

## That is the pipeline

Five generations from vendor exports to a mart. The analyses run on the last
three:

```{r}
analyze_data("clean", ...)     #  6 modules
analyze_data("merged", ...)    # 13 modules
analyze_data("refined", ...)   #  6, the background suite
```
