---
title: "Generation 2: slim"
description: >
  Reshape each pilot database into the shared seven-table schema, then flag
  quality, duplicates, detection limits and derived values.
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

Slim is where the five sources stop looking like five different databases. Every
pilot database, whatever its shape, becomes the same seven tables, and a run of
marking steps then flags what the clean generation will act on.

Nothing is deleted here. The flags are suspicious markers for the next
generation and for manual review, so a slim database has the same measurements
its pilot database had, plus opinions about them.

## Before you start

Only one thing: the source's **pilot database**, `<source>_pilot.sqlite`, in the
same `db_dir`. No raw exports, no seastamp, no `readxl` or `sf` — slim reads
SQLite and writes SQLite, so nothing environmental can bite. If a run fails here
it is the data or the code, which makes this generation a cleaner test than the
pilot was.

## The schema

Seven tables, identical across sources:

```
element → dataset → site → event → subsample → measurement
                                                  ↑
                                               method
```

- `event.dataset_id → dataset`, `event.site_id → site`
- `subsample.event_id → event`
- `measurement.subsample_id → subsample`, `measurement.symbol → element`,
  `measurement.method_id → method`; `method.symbol → element`

The *structure* is identical; the wide tables' columns are not. `measurement`
runs from 15 columns (Mareano) to 24 (ICES-DOME and 4Demon), because each source
carries what it actually recorded: `below_lld` for Mareano, `operator` and
`filtered` for Vannmiljø, `basis`/`matrix`/`qflag` for the rest. The per-source
column map is in [slim-pipeline.md](https://github.com/seafood-hazards/multised-engine/blob/main/docs/slim-pipeline.md).

Sites are keyed on latitude and longitude **rounded to 3 decimal places**, which
is why the site count is lower than the event count everywhere.

## Running it

```{r}
create_db("slim", "mareano", db_dir = "./data/multised-test")
```

Fast — a few seconds per source, since there is no parsing and no external tool.

```{r}
for (src in multised_sources()) create_db("slim", src, db_dir = "./data/multised-test")
```

### Steps

Fifteen numbered steps, but **no source runs all fifteen**. Steps 1-12 are
common; 13 onward exist only where a source has something extra to fold in or
derive, and the numbers are fixed per concern rather than renumbered per source.

```{r}
slim_steps("mareano")   # 1-12, 15
slim_steps("4demon")    # 1-13
```

| Source | Steps | Why |
|---|---|---|
| `mareano` | 1-12, 15 | no source-native flags, grain size already clean |
| `vannmiljo` | 1-15 | all of them |
| `ices-dome` | 1-15 | all of them |
| `mudab` | 1-12, 14, 15 | no source-native flags |
| `4demon` | 1-13 | no grain size, so nothing to correct or derive |

| Step | Adds |
|---|---|
| 1-2 | pilot database → data frames → the seven tables |
| 3 | `element.category` — `target` / `reference` / `organic` / `composition` |
| 4 | `area_flag`, `invalid_flag` |
| 5 | `dup_flag` |
| 6 | the `*_exist` flags on `subsample` |
| 7 | `n_layers`, `multi_flag` |
| 8 | `below_loq` |
| 9 | `value_std`, `unit_std` (mg/kg) |
| 10 | `range_flag` |
| 11 | `below_loq_num` |
| 12 | `weight_basis` |
| 13 | `src_flag` — the source's own flags |
| 14 | `value_std_corr`, `gs_corr` — grain-size correction |
| 15 | `fines_lt63`, `fines_basis` |

Step 3 matters more than its position suggests: `element.category` is the single
source of truth for what a measurand *is*, and every later step reads it rather
than keeping its own list of symbols.

**Steps 1 and 2 are one unit.** The parse builds the frames, the write consumes
them, and neither is useful alone, so asking for either runs both:

```{r}
create_db("slim", "mareano", steps = 5)      # just re-mark duplicates
create_db("slim", "mareano", steps = 2)      # runs 1 and 2
create_db("slim", "mareano", steps = 13)     # error: does not apply to mareano
```

Every step is idempotent, so re-running one, or the whole generation, is safe.

## What you should get

Row counts of the five stored slim databases.

| Source | element | dataset | site | event | subsample | measurement | method |
|---|--:|--:|--:|--:|--:|--:|--:|
| `mareano` | 13 | 2 | 397 | 404 | 3,280 | 41,789 | 23 |
| `vannmiljo` | 20 | 27 | 17,961 | 25,188 | 28,023 | 107,867 | 446 |
| `ices-dome` | 55 | 28 | 6,660 | 13,576 | 14,756 | 158,858 | 1,314 |
| `mudab` | 23 | 16 | 1,838 | 6,033 | 6,209 | 63,642 | 423 |
| `4demon` | 4 | 5 | 160 | 1,480 | 1,480 | 6,739 | 38 |

4Demon has 4 elements because it carries no grain size and no organic carbon;
ICES-DOME has 55 because its grain-size composition vocabulary is large.

Unlike the pilot generation, a slim rebuild should be **byte-identical** to the
stored database — there is no external tool in the path to introduce
differences. A full Mareano rebuild was checked this way: all 7 tables identical,
not merely the same row counts.

## Checking a rebuild

```{r}
compare <- function(stored, rebuilt) {
  a <- DBI::dbConnect(RSQLite::SQLite(), stored)
  b <- DBI::dbConnect(RSQLite::SQLite(), rebuilt)
  on.exit({DBI::dbDisconnect(a); DBI::dbDisconnect(b)}, add = TRUE)
  vapply(DBI::dbListTables(a), function(t)
    identical(DBI::dbReadTable(a, t), DBI::dbReadTable(b, t)), logical(1))
}

compare("./data/db/mareano_slim.sqlite",
        "./data/multised-test/mareano_slim.sqlite")
```

All seven should be `TRUE`. If one is not, compare its dimensions first: a
column difference points at the schema step, a row difference at the transform.

The flag tallies are the other useful check, since they are what the whole
generation exists to produce:

```{r}
con <- DBI::dbConnect(RSQLite::SQLite(), "./data/multised-test/mareano_slim.sqlite")
DBI::dbGetQuery(con, "SELECT category, COUNT(*) n FROM element GROUP BY category")
DBI::dbGetQuery(con, "SELECT dup_flag, COUNT(*) n FROM measurement GROUP BY dup_flag")
DBI::dbGetQuery(con, "SELECT weight_basis, COUNT(*) n FROM measurement GROUP BY weight_basis")
DBI::dbDisconnect(con)
```

## When it fails

**"unable to open database file"** — the pilot database is not in `db_dir`.
Slim reads `<source>_pilot.sqlite` from the same directory it writes to, so
build the pilot generation there first, or copy it across.

**"Step(s) 13 do not apply to mareano"** — the step registry rejecting a step
the source does not have, rather than silently doing nothing. `slim_steps(src)`
lists what applies.

**Row counts that look plausible but are wrong.** This has happened once:
`ices_dome_slim.sqlite` was overwritten with 4Demon output and went unnoticed
because the numbers were believable. Check the source is what you think it is
before trusting a comparison — `SELECT DISTINCT source FROM dataset` is the
quickest tell.

## Next

`create_db("clean", source)` applies the flags this generation set: removing
what is unusable, collapsing replicates, harmonising units and names, and
geo-enriching every site with seastamp.
